### 2017-08-07 paper/paper.tex
bones of a paper

\documentclass[twocolumn]{aastex61}
\usepackage{graphicx}

\usepackage[sort&compress]{natbib}
\usepackage[hang,flushmargin]{footmisc}

\begin{document}
\graphicspath{ {figures/} }
\DeclareGraphicsExtensions{.pdf,.eps,.png}

\title{How to Extract Maximum-Precision Radial Velocities from High-Resolution Spectra}

\author{David W. Hogg}
\affiliation{a lot of places}

\author{Megan Bedell}
\affiliation{Flatiron Institute?}

\begin{abstract}

\end{abstract}

\section{Introduction}

summary of the ways people derive precise RVs from spectra; citations to binary mask papers, papers that compare template matching to binary masks

\section{Overview of Methods}

cross-correlation vs. max likelihood

for either of these, you could use:
\begin{itemize}
\item ``binary'' mask
\item synthetic spectrum
\item data-driven synthetic spectrum
\end{itemize}

information theory based discussion of these: compute the CRLB for velocity

\section{Generating Test Spectra}

how the model works in general, Doppler shifting, parameterization of white noise. number of lines.

additional noise considerations:
\begin{itemize}
\item (micro-) telluric features
\item subtle continuum trends
\item cosmic rays?
\end{itemize}

another important consideration is how well the mask/template fits the data. summarize masks and templates tried (do they know the correct relative line depths and locations?).

\section{Single-Line Results}

\section{Multi-Line Results}

\section{Discussion}

\section{Conclusions}

\acknowledgements

\bibliographystyle{apj}
\bibliography{}%general,myref,inprep}

\end{document}

### 2017-08-07 paper/figures/mkplot_binarymask.py

@@ -1,0 +1,33 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import FormatStrFormatter
+
+if __name__ == "__main__":
+    # load data
+    wave, flux = np.loadtxt('sunspec_500nm.txt', unpack=True, dtype=np.float64)
+    flux /= 1e4
+    wi, wf, weight = np.loadtxt('G2.mas',unpack=True,dtype=np.float64)
+    
+    # setup figure
+    c2 = '#003399' # blue
+    plt.rcParams["font.sans-serif"] = "Helvetica"
+    fig = plt.figure()
+    ax = fig.add_subplot(111)
+    
+    ax.plot(wave, flux, color='black') # plot the Sun
+    
+    # plot mask lines
+    mask = (wi > 4995.) & (wf < 5005.)
+    for wwi,wwf,wweight in zip(wi[mask], wf[mask], weight[mask]):
+        ws = np.arange(wwi,wwf,0.001)
+        fs1 = np.zeros_like(ws)
+        fs2 = np.zeros_like(ws) + np.sqrt(wweight)
+        #fs1 = np.ones_like(ws) - np.sqrt(wweight)
+        #fs2 = np.ones_like(ws)
+        ax.fill_between(ws, fs1, fs2, facecolor=c2, alpha=0.5)
+    
+    # tweak & save    
+    ax.set_xlabel(r'Wavelength $(\AA)$')
+    ax.set_ylabel('Normalized Flux')
+    ax.set_xlim([4995., 5005.])   
+    fig.savefig('binarymask.pdf')

### 2017-08-07 paper/paper.tex
intro text

@@
-\documentclass[twocolumn]{aastex61}
+\documentclass[modern]{aastex61}
 \usepackage{graphicx}
-
+\usepackage{xcolor}
 \usepackage[sort&compress]{natbib}
 \usepackage[hang,flushmargin]{footmisc}
+
+\newcommand{\kms}{km s$^{-1}$}
+\newcommand{\ms}{m s$^{-1}$}
+\newcommand\todo[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
 
@@
 
-summary of the ways people derive precise RVs from spectra; citations to binary mask papers, papers that compare template matching to binary masks
+Accurate spectroscopic measurements of radial velocities (RVs) have long been a critical tool across many subfields of astronomy, from galactic dynamics to cosmic expansion.
+In recent decades, however, a new focus has emerged on \textit{extremely precise} RV measurements of stars as a means of detecting exoplanets.
+For this application, measuring the absolute velocity of the star is much less important than resolving its changes through time.
+With sub-\ms-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
+At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
+In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision RVs: the method of analysis used to extract the RV from a stellar spectrum.
+
+[summary of the ways people derive precise RVs from spectra; citations to binary mask papers, papers that compare template matching to binary masks]
 
@@
 information theory based discussion of these: compute the CRLB for velocity
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{binarymask}
+\caption{Illustration of a quasi-binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
+The vertical length of each mask window scales to the square root of the weight given to the line when performing a cross-correlation between mask and spectrum.}
+\label{fig:binarymask}
+\end{figure}
 

### 2017-08-07 paper/figures/mkplot_binarymask.py

@@ -20,14 +20,15 @@ if __name__ == "__main__":
     mask = (wi > 4995.) & (wf < 5005.)
     for wwi,wwf,wweight in zip(wi[mask], wf[mask], weight[mask]):
         ws = np.arange(wwi,wwf,0.001)
-        fs1 = np.zeros_like(ws)
-        fs2 = np.zeros_like(ws) + np.sqrt(wweight)
-        #fs1 = np.ones_like(ws) - np.sqrt(wweight)
-        #fs2 = np.ones_like(ws)
-        ax.fill_between(ws, fs1, fs2, facecolor=c2, alpha=0.5)
+        #fs1 = np.zeros_like(ws)
+        #fs2 = np.zeros_like(ws) + np.sqrt(wweight)
+        fs1 = np.ones_like(ws) - np.sqrt(wweight)
+        fs2 = np.ones_like(ws)
+        ax.fill_between(ws, fs1, fs2, facecolor=c2, alpha=0.6)
     
     # tweak & save    
     ax.set_xlabel(r'Wavelength $(\AA)$')
     ax.set_ylabel('Normalized Flux')
-    ax.set_xlim([4995., 5005.])   
+    ax.set_xlim([5000., 5005.])
+    ax.set_ylim([0.05, 1.03])   
     fig.savefig('binarymask.pdf')

### 2017-08-07 code/fakedata.py
generalized to four-line case

@@ -25,61 +25,71 @@ def doppler(rv):
 def oned_gaussian(xs, mm, sig):
     return np.exp(-0.5 * (xs - mm) ** 2 / sig ** 2) / np.sqrt(2. * np.pi * sig)
 
-def make_synth(rv, xs, mm, sig):
+def make_synth(rv, xs, ds, ms, sig):
     """
     `rv`: radial velocity in m/s (or same units as `c` above
     `xs`: `[M]` array of wavelength values
-    `mm`: location of the line center in rest wavelength
+    `ds`: depth-like (really EW) parameters for lines
+    `ms`: locations of the line centers in rest wavelength
     `sig`: sigma of pixel-convolved LSF
     """
     synths = np.ones_like(xs)
-    synths -= oned_gaussian(xs * doppler(rv), mm, sig) / 8.
+    for d, m in zip(ds, ms):
+        synths *= np.exp(-d *
+            oned_gaussian(xs * doppler(rv), m, sig))
     return synths
 
-def make_mask(rv, xs, mm, w1, w2):
+def make_mask(rv, xs, ws, ms, w1, w2):
     """
     `rv`: radial velocity in m/s (or same units as `c` above
     `xs`: [M] array of wavelength values
-    `mm`: location of the center of the binary mask tophat in rest wavelength
+    `ws`: weights to apply to the binary mask tophats
+    `ms`: locations of the centers of the binary mask tophats in rest wavelength
     `w1`: half-width of a wavelength pixel
     `w2`: half-width of the binary mask tophat in rest wavelength
 
     Notes: The doppler "shift" dilates the input wavelengths,
       and also the w1 half-width.
+    
+    Bugs: Super-inefficient code.
     """
     assert w1 > 0
     assert w2 > 0
     drv = doppler(rv)
     ddd = np.abs(w1 * drv - w2)
     dd = w1 * drv + w2
-    xmms = xs * drv - mm
     model = np.zeros_like(xs)
-    model[xmms > -dd] = (dd + xmms[xmms > -dd]) / (dd - ddd)
-    model[xmms > -ddd] = 1.
-    model[xmms > ddd] = (dd - xmms[xmms > ddd]) / (dd - ddd)
-    model[xmms > dd] = 0.
+    for ww, mm in zip(ws, ms):
+        xmms = xs * drv - mm
+        dmodel = np.zeros_like(xs)
[…]

### 2017-08-08 paper/paper.tex
more realistic fake spectra; template made by stacking epochs

@@
 \item ``binary'' mask
-\item synthetic spectrum
-\item data-driven synthetic spectrum
+\item synthetic spectrum (or template observation?)
+\item data-driven template spectrum
 \end{itemize}
@@
 
-\section{Generating Test Spectra}
+\section{Testing with Toy-Model Spectra}
 
-how the model works in general, Doppler shifting, parameterization of white noise. number of lines.
+The basic model that we use to generate fake spectra is extremely simple, assuming a perfectly normalized continuum; isolated, perfectly Gaussian spectral lines; and white noise only.
+We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single RV spectrum.
+Lines were inserted as Gaussians with arbitrarily-prescribed centers and depths (or equivalent widths).
+Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 $\AA$.
+%The number of lines used varies in the results discussed below.
+
+We repeated this synthesis to make a set of 512 spectra, each with a random RV between -30 \kms and 30 \kms (the approximate amplitude of the yearly RV shifts induced by barycentric motion).
+The RVs can be applied as a straightforward Doppler shift in the centers of each line.
+\todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered RVs from the input RVs.
+
+\todo{remember to discuss quadratic max procedure.. probably in previous section?}
+
+\subsection{Single-Line Results}
+
+\subsection{Multi-Line Results}
+
+\section{Dependence of RV Precision on Template Quality}
+
+how well does the mask/template approximate the real data? try fudging the relative line depths and locations.
+
+\section{Correlated Noise Sources and Their Consequences}
 
@@
 
-another important consideration is how well the mask/template fits the data. summarize masks and templates tried (do they know the correct relative line depths and locations?).
+\section{Application to Real Spectra}
 
-\section{Single-Line Results}
-
-\section{Multi-Line Results}
+how do our results compare to the reality of the situation? (will this tell us anything other than ``real spectra are more complicated and our template is silly''?)
 

### 2017-08-08 code/fakedata.py

@@ -15,6 +15,7 @@ Copyright 2017 David W Hogg (NYU).
 """
 
 import numpy as np
+from scipy.interpolate import interp1d
 
 c = 299792458. # m/s
 
@@ -25,17 +26,17 @@ def doppler(rv):
 def oned_gaussian(xs, mm, sig):
     return np.exp(-0.5 * (xs - mm) ** 2 / sig ** 2) / np.sqrt(2. * np.pi * sig)
 
-def make_synth(rv, xs, ds, ms, sig):
+def make_synth(rv, xs, ds, ms, sigs):
     """
     `rv`: radial velocity in m/s (or same units as `c` above
     `xs`: `[M]` array of wavelength values
-    `ds`: depth-like (really EW) parameters for lines
+    `ds`: depths at line centers
     `ms`: locations of the line centers in rest wavelength
-    `sig`: sigma of pixel-convolved LSF
+    `sigs`: Gaussian sigmas of lines
     """
     synths = np.ones_like(xs)
-    for d, m in zip(ds, ms):
-        synths *= np.exp(-d *
+    for d, m, sig in zip(ds, ms, sigs):
+        synths *= np.exp(d *
             oned_gaussian(xs * doppler(rv), m, sig))
     return synths
 
@@ -68,20 +69,51 @@ def make_mask(rv, xs, ws, ms, w1, w2):
         dmodel[xmms > dd] = 0.
         model += ww * dmodel
     return 1. - model
+    
+def make_template(all_data, rvs, xs, dx, plot=False, plotname='template.png'):
+    """
+    `all_data`: `[N, M]` array of pixels
+    `rvs`: `[N]` array of RVs
+    `xs`: `[M]` array of wavelength values
+    `dx`: linear spacing desired for template wavelength grid (A)
+    """
+    (N,M) = np.shape(all_data)
+    all_xs = np.empty_like(all_data)
+    for i in range(N):
+        all_xs[i,:] = xs * doppler(rvs[i]) # shift to rest frame
+    all_data, all_xs = np.ravel(all_data), np.ravel(all_xs)
+    template_xs = np.arange(min(all_xs), max(all_xs), dx)
+    template_ys = np.empty_like(template_xs)
+    for i,t in enumerate(template_xs):
+        ind = (all_xs >= t-dx/2.) & (all_xs <= t+dx/2.)
+        template_ys[i] = np.median(all_data[ind])
+    if plot == True:
[…]

### 2017-08-08 paper/figures/mkplot_binarymask.py

@@ -17,7 +17,7 @@ if __name__ == "__main__":
     ax.plot(wave, flux, color='black') # plot the Sun
     
     # plot mask lines
-    mask = (wi > 4995.) & (wf < 5005.)
+    mask = (wi > 4997.8) & (wf < 5002.2)
     for wwi,wwf,wweight in zip(wi[mask], wf[mask], weight[mask]):
         ws = np.arange(wwi,wwf,0.001)
         #fs1 = np.zeros_like(ws)
@@ -29,6 +29,6 @@ if __name__ == "__main__":
     # tweak & save    
     ax.set_xlabel(r'Wavelength $(\AA)$')
     ax.set_ylabel('Normalized Flux')
-    ax.set_xlim([5000., 5005.])
+    ax.set_xlim([4997.8, 5002.2])
     ax.set_ylim([0.05, 1.03])   
     fig.savefig('binarymask.pdf')

### 2017-08-10 code/fakedata.py
iterative template-building

@@ -77,6 +77,7 @@ def make_template(all_data, rvs, xs, dx, plot=False, plotname='template.png'):
     `xs`: `[M]` array of wavelength values
     `dx`: linear spacing desired for template wavelength grid (A)
     """
+    foo = 30.
     (N,M) = np.shape(all_data)
     all_xs = np.empty_like(all_data)
     for i in range(N):
@@ -85,11 +86,11 @@ def make_template(all_data, rvs, xs, dx, plot=False, plotname='template.png'):
     template_xs = np.arange(min(all_xs), max(all_xs), dx)
     template_ys = np.empty_like(template_xs)
     for i,t in enumerate(template_xs):
-        ind = (all_xs >= t-dx/2.) & (all_xs <= t+dx/2.)
-        template_ys[i] = np.median(all_data[ind])
+        ind = (all_xs >= t-dx/2.) & (all_xs < t+dx/2.)
+        template_ys[i] = np.sum(all_data[ind]) / np.sum(ind)
     if plot == True:
         plt.clf()
-        plt.scatter(all_xs, all_data)
+        plt.scatter(all_xs, all_data, marker=".", alpha=0.25)
         plt.plot(template_xs, template_ys, color='black', lw=2)
         plt.title('Fitting a template to all data')
         plt.savefig(plotname)
@@ -97,7 +98,7 @@ def make_template(all_data, rvs, xs, dx, plot=False, plotname='template.png'):
     
 def shift_template(rv, xs, template_xs, template_ys):
     f = interp1d(template_xs / doppler(rv), template_ys, bounds_error = False, 
-            fill_value = np.nan)
+            fill_value = 1.)
     return f(xs)
     
 
@@ -213,7 +214,7 @@ if __name__ == "__main__":
         plotprefix = "realistic"
 
     # make and plot fake data
-    N = 512
+    N = 16
     data, ivars, true_rvs = make_data(N, xs, ds, ms, sigs)
     plt.clf()
     for n in range(8):
@@ -232,91 +233,146 @@ if __name__ == "__main__":
         crlbs[n] = np.sum(dmodel_dv * ivars[n, :] * dmodel_dv)
     crlb = 1. / np.sqrt(np.mean(crlbs))
     print("CRLB:", crlb, "m/s")
-
-    # get best-fit velocities
-    halfwidth = 0.075 # A; half-width of binary mask
-    options = ([make_synth, (xs, ds, ms, sigs), xcorr],
[…]

### 2017-08-10 code/fakedata.py
continuum normalization

@@ -16,6 +16,7 @@ Copyright 2017 David W Hogg (NYU).
 
 import numpy as np
 from scipy.interpolate import interp1d
+from scipy.optimize import curve_fit
 
 c = 299792458. # m/s
 
@@ -125,6 +126,35 @@ def make_data(N, xs, ds, ms, sigs):
         data[n, :] = make_synth(rv, xs, ds, ms, sigs)
         data[n, :] += np.random.normal(size=M) / np.sqrt(ivars[n, :])
     return data, ivars, rvs
+    
+def quadratic(xs, p0, p1, p2):
+    return p0 + p1*xs + p2*xs**2
+    
+def continuum_normalize(xs, data, ivars, percents=(70., 99.), plot=False, plotname='continuum_normalization.png'):
+    """
+    `xs`: `[M]` array of wavelength values
+    `data`: `[M]` array of pixels
+    `ivars`: `[M]` array of inverse variance values
+
+    Note: Presumes `ivars` is a list of diagonal entries; no
+      capacity for a dense inverse-variance matrix.
+    """
+    (lo, hi) = np.percentile(data, percents)
+    ind_fit = (data >= lo) & (data <= hi)
+    popt, pcov = curve_fit(quadratic, xs[ind_fit], data[ind_fit], 
+                    sigma=np.sqrt(1./ivars[ind_fit]), absolute_sigma=True)
+    continuum = quadratic(xs, popt[0], popt[1], popt[2])
+    normalized_data = data / continuum
+    normalized_ivars = ivars * continuum**2
+    if plot:
+        plt.clf()
+        plt.scatter(xs, data, color='k', alpha=0.5)
+        plt.scatter(xs[ind_fit], data[ind_fit], color='red')
+        plt.plot(xs, continuum)
+        plt.title('continuum normalization')
+        plt.savefig(plotname)
+    return normalized_data, normalized_ivars
+    
 
 def xcorr(data, model, ivars):
     """
@@ -221,6 +251,14 @@ if __name__ == "__main__":
         plt.step(xs, data[n, :] + 0.25 * n, color="k")
     plt.title("examples of (fake) data")
     plt.savefig(plotprefix+"_data.png")
+    
+    # continuum normalize
+    for n in range(N):
+        if n < 4:
+            data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :], plot=True,
+                plotname=plotprefix+"_normalization{0}.png".format(n))
+        else:
[…]

### 2017-08-10 code/fakedata.py
added tellurics

@@ -127,10 +127,23 @@ def make_data(N, xs, ds, ms, sigs):
         data[n, :] += np.random.normal(size=M) / np.sqrt(ivars[n, :])
     return data, ivars, rvs
     
+def add_tellurics(xs, all_data, true_rvs, lambdas, strengths, dx, plot=False, plotname='tellurics.png'):
+    N, M = np.shape(all_data)
+    tellurics = np.ones_like(xs)
+    for ll, s in zip(lambdas, strengths):
+        tellurics *= np.exp(-s * oned_gaussian(xs, ll, dx))
+    all_data *= np.repeat([tellurics,],N,axis=0)
+    if plot:
+        plt.clf()
+        plt.plot(xs, tellurics)
+        plt.title('tellurics model')
+        plt.savefig(plotname)
+    return all_data
+    
 def quadratic(xs, p0, p1, p2):
     return p0 + p1*xs + p2*xs**2
     
-def continuum_normalize(xs, data, ivars, percents=(70., 99.), plot=False, plotname='continuum_normalization.png'):
+def continuum_normalize(xs, data, ivars, percents=(80., 99.), plot=False, plotname='continuum_normalization.png'):
     """
     `xs`: `[M]` array of wavelength values
     `data`: `[M]` array of pixels
@@ -243,9 +256,20 @@ if __name__ == "__main__":
         xs = np.arange(4998. + 0.5 * dx, 5002., dx) # A
         plotprefix = "realistic"
 
-    # make and plot fake data
-    N = 16
+    # make fake data
+    N = 128
     data, ivars, true_rvs = make_data(N, xs, ds, ms, sigs)
+    
+    # add tellurics
+    if True:
+        n_tellurics = 8
+        np.random.seed(75)
+        lambdas = np.random.uniform(xs[0], xs[-1], n_tellurics)
+        strengths = 1. - np.random.power(700., n_tellurics)
+        data = add_tellurics(xs, data, true_rvs, lambdas, strengths, dx, plot=True)
+        np.random.seed(42)
+        
+    # plot the data    
     plt.clf()
     for n in range(8):
         plt.step(xs, data[n, :] + 0.25 * n, color="k")
@@ -272,6 +296,7 @@ if __name__ == "__main__":
     crlb = 1. / np.sqrt(np.mean(crlbs))
     print("CRLB:", crlb, "m/s")
     
+    
[…]

### 2017-08-11 code/fakedata.py
exported from main to functions, added real data capabilities

@@ -3,7 +3,6 @@ This code is part of the EPRV project.
 Copyright 2017 David W Hogg (NYU).
 
 ## To-do:
-- Make EW of line an input variable in make_synth and propagate.
 - Optimize width of binary mask window.
 -- Do this with a function that is general-purpose
 - Look at errors as a function of synthetic-spectrum wrongness
@@ -11,12 +10,13 @@ Copyright 2017 David W Hogg (NYU).
 -- in width
 -- in line shape
 - Consider continuum noise or wrongness
-- Make a model with 2 lines or 4 to test scalings.
 """
 
 import numpy as np
 from scipy.interpolate import interp1d
 from scipy.optimize import curve_fit
+import pylab as plt
+
 
 c = 299792458. # m/s
 
@@ -41,14 +41,14 @@ def make_synth(rv, xs, ds, ms, sigs):
             oned_gaussian(xs * doppler(rv), m, sig))
     return synths
 
-def make_mask(rv, xs, ws, ms, w1, w2):
+def make_mask(rv, xs, ws, ms, w1, w2s):
     """
     `rv`: radial velocity in m/s (or same units as `c` above
     `xs`: [M] array of wavelength values
     `ws`: weights to apply to the binary mask tophats
     `ms`: locations of the centers of the binary mask tophats in rest wavelength
     `w1`: half-width of a wavelength pixel
-    `w2`: half-width of the binary mask tophat in rest wavelength
+    `w2s`: half-widths of the binary mask tophat in rest wavelength
 
     Notes: The doppler "shift" dilates the input wavelengths,
       and also the w1 half-width.
@@ -56,12 +56,12 @@ def make_mask(rv, xs, ws, ms, w1, w2):
     Bugs: Super-inefficient code.
     """
     assert w1 > 0
-    assert w2 > 0
+    #assert w2 > 0
     drv = doppler(rv)
-    ddd = np.abs(w1 * drv - w2)
-    dd = w1 * drv + w2
     model = np.zeros_like(xs)
-    for ww, mm in zip(ws, ms):
+    for ww, mm, w2 in zip(ws, ms, w2s):
+        ddd = np.abs(w1 * drv - w2)
+        dd = w1 * drv + w2
         xmms = xs * drv - mm
         dmodel = np.zeros_like(xs)
[…]

### 2017-08-11 code/realdata.py

@@ -1,0 +1,111 @@
+"""
+This code is part of the EPRV project.
+Copyright 2017 David W Hogg (NYU), Megan Bedell (Flatiron, UChicago).
+
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from fakedata import *
+
+c = 299792458. # m/s
+
+if __name__ == "__main__":
+    harps_mask = True # mask choice
+    
+    # load the data for quiet star HIP54287 (HARPS RMS 1.3 m/s)
+    # BUGS: directories and dependencies will only work on Megan's computer...
+    from scipy.io.idl import readsav
+    from wobble import read_harps, rv_model    
+    data_dir = "/Users/mbedell/Documents/Research/HARPSTwins/Results/"
+    s = readsav(data_dir+'HIP54287_result.dat')
+    Star = rv_model.RV_Model()
+    N = len(s.files)  # number of epochs
+    data = np.empty((N, 400))
+    ivars = np.empty_like(data)
+    dx = 0.01 # A
+    xs = np.arange(4998., 5002., dx)
+    for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
+        # read in the spectrum
+        spec_file = str.replace(f, 'ccf_G2', 's1d')
+        wave, spec = read_harps.read_spec(spec_file)
+        # re-introduce barycentric velocity
+        wave /= doppler(b*1.e3)
+        # remove systemic RV shift so we're looking at the same lines as example
+        wave *= doppler(54.9 * 1.e3)
+        # save the relevant bit
+        f = interp1d(wave, spec)
+        data[n,:] = f(xs)
+        ivars[n,:] = snr**2
+    true_rvs = (s.berv + s.rv - 54.9) * 1.e3  # m/s
+    plotprefix = 'harpsdata'
+    
+    # continuum normalize
+    for n in range(N):
+        if n < 4:
+            data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :], plot=True,
+                plotname=plotprefix+"_normalization{0}.png".format(n), percents=(60,99))
+        else:
+            data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :])
+    
+    # plot the data    
+    plot_data(xs, data, tellurics=False, plotname=plotprefix+"data.png")
+    
+    
+    # make a perfect template from stacked observations
[…]

### 2017-08-11 paper/paper.tex
expanded to a whole order

@@
 
-\newcommand{\kms}{km s$^{-1}$}
-\newcommand{\ms}{m s$^{-1}$}
+\newcommand{\kms}{km\,s$^{-1}$}
+\newcommand{\ms}{m\,s$^{-1}$}
 \newcommand\todo[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
@@
 
+\author{Megan Bedell}
+\affiliation{Flatiron Institute}
+\affiliation{University of Chicago}
+
 \author{David W. Hogg}
-\affiliation{a lot of places}
-
-\author{Megan Bedell}
-\affiliation{Flatiron Institute?}
+\affiliation{Flatiron Institute}
+\affiliation{CCPP, NYU}
+\affiliation{CDS, NYU}
+\affiliation{MPIA}
 
 \begin{abstract}
-
+Precise stellar radial velocity (RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
+Despite the importance of these observations, the detailed computation methods used to extract \ms-level RVs from stellar spectra remain largely unstandardized and unpublished.
+In this work, we provide a brief overview of possible approaches to extracting RVs and compare their efficiencies by testing on a segment of realistic model spectrum.
+We find that cross-correlation with an iteratively built data-driven template outperforms traditional cross-correlation with a quasi-binary mask by an order of magnitude and saturates the Cram\'{e}r-Rao bound under white-noise conditions.
+We additionally simulate the effects of unresolved micro-telluric lines and quantify their influence on the achievable RV precision.
+We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the RV precision of HARPS to substantially better than 1 \ms.
 \end{abstract}
@@
 \includegraphics[width=\columnwidth]{binarymask}
-\caption{Illustration of a quasi-binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
+\caption{Illustration of a weighted binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
 The vertical length of each mask window scales to the square root of the weight given to the line when performing a cross-correlation between mask and spectrum.}
@@
 \acknowledgements
+MPIA hospitality
 

### 2017-08-11 code/realdata.py

@@ -20,11 +20,12 @@ if __name__ == "__main__":
     data_dir = "/Users/mbedell/Documents/Research/HARPSTwins/Results/"
     s = readsav(data_dir+'HIP54287_result.dat')
     Star = rv_model.RV_Model()
-    N = len(s.files)  # number of epochs
-    data = np.empty((N, 400))
-    ivars = np.empty_like(data)
     dx = 0.01 # A
-    xs = np.arange(4998., 5002., dx)
+    xs = np.arange(4987.4, 5043.5, dx)
+    N = len(s.files)  # number of epochs
+    M = len(xs)
+    data = np.empty((N, M))
+    ivars = np.empty_like(data)
     for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
         # read in the spectrum
         spec_file = str.replace(f, 'ccf_G2', 's1d')
@@ -44,12 +45,12 @@ if __name__ == "__main__":
     for n in range(N):
         if n < 4:
             data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :], plot=True,
-                plotname=plotprefix+"_normalization{0}.png".format(n), percents=(60,99))
+                plotname=plotprefix+"_normalization{0}.png".format(n), percents=(90,99))
         else:
             data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :])
     
     # plot the data    
-    plot_data(xs, data, tellurics=False, plotname=plotprefix+"data.png")
+    plot_data(xs, data, tellurics=False, plotname=plotprefix+"_data.png")
     
     
     # make a perfect template from stacked observations

### 2017-09-11 paper/paper.tex
syncing old changes, full order fitting

@@
 
-\title{How to Extract Maximum-Precision Radial Velocities from High-Resolution Spectra}
+\title{Extracting Maximum-Precision Radial Velocities from High-Resolution Spectra}
 
@@
 With sub-\ms-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
-At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
+At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
 In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision RVs: the method of analysis used to extract the RV from a stellar spectrum.
+
+While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of template
+
+We begin in Section \ref{s:methods} with a summary of the practices used by current RV pipelines, including cross-correlation and template fitting.
+We then implement these techniques to extract RVs from a 5-\r{A} section of simulated spectrum and compare their performance in Section \ref{s:fakedata}.
 
@@
 \section{Overview of Methods}
+\label{s:methods}
 
@@
 \caption{Illustration of a weighted binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
-The vertical length of each mask window scales to the square root of the weight given to the line when performing a cross-correlation between mask and spectrum.}
+The vertical length of each mask window corresponds to the weight given to the line when performing a cross-correlation between mask and spectrum.}
 \label{fig:binarymask}

### 2017-09-11 code/fakedata.py

@@ -307,7 +307,7 @@ def plot_data(xs, data, tellurics=False, telluric_xs=[0.0], plotname="data.png")
             assert telluric_xs.all() > 0.0
             for t in telluric_xs:
                 plt.axvline(t, color='b', alpha=0.25, lw=1)
-    plt.title("examples of (fake) data")
+    plt.title("examples of data")
     plt.savefig(plotname)
     
 def plot_resids(rvs, true_rvs, crlb=0.0, title='', plotname='rv_mistakes.png'):
@@ -358,7 +358,8 @@ if __name__ == "__main__":
     data, ivars, true_rvs = make_data(N, xs, ds, ms, sigs)
     
     # add tellurics
-    if True:
+    tellurics = True
+    if tellurics:
         n_tellurics = 16 # magic
         telluric_sig = 0.1 # magic
         telluric_xs = np.random.uniform(xs[0], xs[-1], n_tellurics)
@@ -374,7 +375,11 @@ if __name__ == "__main__":
         else:
             data[n, :], ivars[n, :] = continuum_normalize(xs, data[n, :], ivars[n, :])
             
-    plot_data(xs, data, tellurics=True, telluric_xs=telluric_xs, plotname=plotprefix+"data.png")
+    if tellurics:
+        plot_data(xs, data, tellurics=True, telluric_xs=telluric_xs, plotname=plotprefix+"data.png")
+    else:
+        plot_data(xs, data, plotname=plotprefix+"data.png")
+        
     
     # make a perfect template from stacked observations
     template_xs, template_ys = make_template(data, true_rvs, xs, dx, plot=True, 
@@ -425,7 +430,7 @@ if __name__ == "__main__":
       
 
 
-    if False:
+    if True:
         # comparative tests of methods        
         # get best-fit velocities
         halfwidth = 0.075 # A; half-width of binary mask

### 2017-09-11 code/realdata.py

@@ -20,32 +20,74 @@ if __name__ == "__main__":
     data_dir = "/Users/mbedell/Documents/Research/HARPSTwins/Results/"
     s = readsav(data_dir+'HIP54287_result.dat')
     Star = rv_model.RV_Model()
-    dx = 0.01 # A
-    xs = np.arange(4987.4, 5043.5, dx)
-    N = len(s.files)  # number of epochs
-    M = len(xs)
-    data = np.empty((N, M))
-    ivars = np.empty_like(data)
-    for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
-        # read in the spectrum
-        spec_file = str.replace(f, 'ccf_G2', 's1d')
-        wave, spec = read_harps.read_spec(spec_file)
-        # re-introduce barycentric velocity
-        wave /= doppler(b*1.e3)
-        # remove systemic RV shift so we're looking at the same lines as example
-        wave *= doppler(54.9 * 1.e3)
-        # save the relevant bit
-        f = interp1d(wave, spec)
-        data[n,:] = f(xs)
-        ivars[n,:] = snr**2
-    true_rvs = (s.berv + s.rv - 54.9) * 1.e3  # m/s
-    plotprefix = 'harpsdata'
+    
+    if False:
+        true_rvs = (s.berv + s.rv - 54.9) * 1.e3  # m/s
+        drift = s.drift # m/s
+        dx = 0.01 # A
+        xs = np.arange(4987.4, 5043.5, dx)
+        N = len(s.files)  # number of epochs
+        M = len(xs)
+        data = np.empty((N, M))
+        ivars = np.empty_like(data)
+        for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
+            # read in the spectrum
+            spec_file = str.replace(f, 'ccf_G2', 's1d')
+            wave, spec = read_harps.read_spec(spec_file)
+            # re-introduce barycentric velocity
+            wave /= doppler(b*1.e3)
+            # remove systemic RV shift so we're looking at the same lines as example
+            wave *= doppler(54.9 * 1.e3)
+            # save the relevant bit
+            f = interp1d(wave, spec)
+            data[n,:] = f(xs)
+            ivars[n,:] = snr**2
+        plotprefix = 'harpsdata'
+    
+    if True:
+        true_rvs = (s.berv + s.rv - 54.9) * 1.e3  # m/s
[…]

### 2017-09-11 paper/figures/mkplot_binarymask.py

@@ -21,8 +21,8 @@ if __name__ == "__main__":
     for wwi,wwf,wweight in zip(wi[mask], wf[mask], weight[mask]):
         ws = np.arange(wwi,wwf,0.001)
         #fs1 = np.zeros_like(ws)
-        #fs2 = np.zeros_like(ws) + np.sqrt(wweight)
-        fs1 = np.ones_like(ws) - np.sqrt(wweight)
+        #fs2 = np.zeros_like(ws) + wweight
+        fs1 = np.ones_like(ws) - wweight
         fs2 = np.ones_like(ws)
         ax.fill_between(ws, fs1, fs2, facecolor=c2, alpha=0.6)
     

### 2017-11-11 paper/paper.tex
comment out
improved abstract and typography

--- 2017-08-08 paper/paper.tex
@@
 
-\newcommand{\kms}{km s$^{-1}$}
-\newcommand{\ms}{m s$^{-1}$}
+\newcommand{\unit}[1]{\mathrm{#1}}
+\newcommand{\km}{\unit{km}}
+\newcommand{\m}{\unit{m}}
+\newcommand{\s}{\unit{s}}
+\newcommand{\kms}{\km\,\s^{-1}}
+\newcommand{\ms}{\m\,\s^{-1}}
 \newcommand\todo[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
+\newcommand{\acronym}[1]{{\small{#1}}}
+\newcommand{\project}[1]{\textsl{#1}}
+\newcommand{\HARPS}{\project{\acronym{HARPS}}}
+\newcommand{\RV}{\acronym{RV}}
+\newcommand{\CRLB}{\acronym{CRLB}}
 
@@
 
-\title{How to Extract Maximum-Precision Radial Velocities from High-Resolution Spectra}
-
-\author{David W. Hogg}
-\affiliation{a lot of places}
+\title{Achieving maximum possible precision on stellar radial-velocity measurements in multi-epoch spectroscopy}
 
 \author{Megan Bedell}
-\affiliation{Flatiron Institute?}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
+\affiliation{University of Chicago, NEED DETAILS HERE}
+
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
+\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
+\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}
 
 \begin{abstract}
-
+% context
+Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
+Despite the importance of these observations, the detailed computation methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
+% aims
+In this work, we look at sensible approaches to extracting \RV s from a data set of  one-dimensional, calibrated, extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
+% methods
+We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
+We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
+% results
+We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
+We find that if there are micro-tellurics, they can be captured by a data-driven model.
+We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
 \end{abstract}
@@
 
-Accurate spectroscopic measurements of radial velocities (RVs) have long been a critical tool across many subfields of astronomy, from galactic dynamics to cosmic expansion.
-In recent decades, however, a new focus has emerged on \textit{extremely precise} RV measurements of stars as a means of detecting exoplanets.
+Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from galactic dynamics to cosmic expansion.
+In recent decades, however, a new focus has emerged on \textit{extremely precise} \RV\ measurements of stars as a means of detecting exoplanets.
 For this application, measuring the absolute velocity of the star is much less important than resolving its changes through time.
-With sub-\ms-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
-At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
-In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision RVs: the method of analysis used to extract the RV from a stellar spectrum.
+With sub-$\ms$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
+At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
+In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision \RV s: the method of analysis used to extract the \RV\ from a stellar spectrum.
 
-[summary of the ways people derive precise RVs from spectra; citations to binary mask papers, papers that compare template matching to binary masks]
+While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of template
+
+We begin in Section \ref{s:methods} with a summary of the practices used by current \RV\ pipelines, including cross-correlation and template fitting.
+We then implement these techniques to extract \RV s from a 5-\r{A} section of simulated spectrum and compare their performance in Section \ref{s:fakedata}.
+
+[summary of the ways people derive precise \RV s from spectra; citations to binary mask papers, papers that compare template matching to binary masks]
 
 \section{Overview of Methods}
+\label{s:methods}
 
@@
 \centering
-\includegraphics[width=\columnwidth]{binarymask}
-\caption{Illustration of a quasi-binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
-The vertical length of each mask window scales to the square root of the weight given to the line when performing a cross-correlation between mask and spectrum.}
+% \includegraphics[width=\columnwidth]{binarymask}
+\caption{Illustration of a weighted binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
+The vertical length of each mask window corresponds to the weight given to the line when performing a cross-correlation between mask and spectrum.}
 \label{fig:binarymask}
@@
 The basic model that we use to generate fake spectra is extremely simple, assuming a perfectly normalized continuum; isolated, perfectly Gaussian spectral lines; and white noise only.
-We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single RV spectrum.
+We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single \RV\ spectrum.
 Lines were inserted as Gaussians with arbitrarily-prescribed centers and depths (or equivalent widths).
@@
 
-We repeated this synthesis to make a set of 512 spectra, each with a random RV between -30 \kms and 30 \kms (the approximate amplitude of the yearly RV shifts induced by barycentric motion).
-The RVs can be applied as a straightforward Doppler shift in the centers of each line.
-\todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered RVs from the input RVs.
+We repeated this synthesis to make a set of 512 spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion).
+The \RV s can be applied as a straightforward Doppler shift in the centers of each line.
+\todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.
 

### 2017-11-27 paper/paper.tex
tidying up

--- 2017-09-11 paper/paper.tex
@@
 
-information theory based discussion of these: compute the CRLB for velocity
+information theory based discussion of these: compute the CRLB for velocity. Mention how uncertainties are typically computed (Butler 1996 and Bouchy 2001 formulae).
 
@@
 We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single RV spectrum.
-Lines were inserted as Gaussians with arbitrarily-prescribed centers and depths (or equivalent widths).
+Lines were inserted as Gaussians with arbitrarily-prescribed centers and depths.
 Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 $\AA$.
-%The number of lines used varies in the results discussed below.
 
 We repeated this synthesis to make a set of 512 spectra, each with a random RV between -30 \kms and 30 \kms (the approximate amplitude of the yearly RV shifts induced by barycentric motion).
-The RVs can be applied as a straightforward Doppler shift in the centers of each line.
+The RVs can be applied as a straightforward Doppler shift on the central wavelength of each line.
 \todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered RVs from the input RVs.
@@
 
+We began with the simplest test case possible: a single, strong absorption feature.
+
+What is the CRLB?
+
+Discuss scaling of the CRLB with line FWHM, constrast/depth, etc.
+According to Lovis \& Fischer (2010) and Figueira (2017), $\sigma_{RV}$ should go as $\frac{\sqrt{FWHM}}{SNR \times C}$, where $C$ is the line contrast, or the depth of the line as a fraction of the continuum level \todo{(check on this definition of contrast)}.
+Do we see this?
+
 \subsection{Multi-Line Results}
 
-\section{Dependence of RV Precision on Template Quality}
+Our next test involved a more realistic test spectrum.
+We measured the centers and depths of \todo{N} lines in the Solar spectrum between {x - y} $\AA$ and used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
 
-how well does the mask/template approximate the real data? try fudging the relative line depths and locations.
+What is the CRLB for this case?
+Binary mask cross-correlation is very far from saturating it.
+Cross-correlation or max-likelihood fitting with a perfect spectral template works much better.
+Talk a little bit about why.
+
+\section{Dependence of RV Precision on Template Choice}
+
+how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template.
+
+is this where we bring in the data-driven template and demonstrate how well it performs?
 
@@
 
-\section{Discussion}
+%\section{Discussion}
 

### 2017-11-27 code/fakedata.py

@@ -107,7 +107,7 @@ def make_template(all_data, rvs, xs, dx, plot=False, plotname='template.png'):
     `xs`: `[M]` array of wavelength values
     `dx`: linear spacing desired for template wavelength grid (A)
     """
-    foo = 30.
+    foo = 30. #magic
     (N,M) = np.shape(all_data)
     all_xs = np.empty_like(all_data)
     for i in range(N):

### 2017-11-27 code/realdata.py

@@ -77,6 +77,34 @@ if __name__ == "__main__":
         #true_rvs = (s.rv - 54.9) * 1.e3 
         drift = s.drift # m/s
         dx = 0.01 # A
+        xs = np.arange(4998.0, 5002.0, dx)
+        N = len(s.files)  # number of epochs
+        M = len(xs)
+        data = np.empty((N, M))
+        ivars = np.empty_like(data)
+        for n,(f,b,snr) in enumerate(zip(s.files, s.berv, s.snr)):
+            # read in the spectrum
+            spec_file = str.replace(f, 'ccf_G2', 's1d')
+            wave, spec = read_harps.read_spec(spec_file)
+            # re-introduce barycentric velocity
+            wave *= doppler(b*1.e3)
+            # remove systemic RV shift so we're looking at the same lines as example
+            wave *= doppler(54.9 * 1.e3)
+            # save the relevant bit
+            f = interp1d(wave, spec)
+            data[n,:] = f(xs)
+            ivars[n,:] = snr**2
+            
+        p0 = None # starting guess for continuum normalization
+        iterate = 1
+        plotprefix = 'harpsdata'
+    
+    if False:
+        # grab a tellurics section from s1d files
+        true_rvs = ( -s.berv + s.rv - 54.9) * 1.e3  # m/s
+        #true_rvs = (s.rv - 54.9) * 1.e3 
+        drift = s.drift # m/s
+        dx = 0.01 # A
         #xs = np.arange(4998.0, 5002.0, dx)
         xs = np.arange(5910.0, 5925.0, dx) # tellurics region
         N = len(s.files)  # number of epochs
@@ -182,7 +210,7 @@ if __name__ == "__main__":
     template_xs, template_ys = make_template(data, true_rvs, xs, dx, plot=True, 
                     plotname=plotprefix+'_perfecttemplate.png')
                     
-    if True:
+    if False:
         # tellurics experiment
         
         # adopt CRLB from fake data experiment
@@ -295,7 +323,7 @@ if __name__ == "__main__":
            
 
                     
-    if False: 
+    if True: 
         # try to get RVs
         
         # adopt CRLB from fake data experiment

### 2017-11-28 paper/paper.tex
outline

@@
 Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
-Despite the importance of these observations, the detailed computation methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
+Despite the importance of these observations, the detailed computational methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
 % aims
-In this work, we look at sensible approaches to extracting \RV s from a data set of  one-dimensional, calibrated, extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
+In this work, we look at sensible approaches to extracting \RV s from a data set of extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
 % methods
@@
 
+The assumption of this project is that we have many observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
+This means we are not considering the case of a real-time data reduction pipeline.
+We also assume a HARPS-like setup, i.e.
+no absorption cell, but the points we make can be easily generalized to the gas-cell case.
+We return to this point in the discussion \todo{(do this)}.
+
+We additionally assume that the data are extremely well wavelength-calibrated.
+In other work, we will discuss this, but our current belief is that current RV spectrographs are supremely well wavelength-calibrated.
+
+\section{Information Theory}
+\label{s:info}
+
+\begin{itemize}
+\item the Cramer-Rao bound is the limit on measurement
+\item it relates to Fisher information
+\item it cannot be computed without a model
+\item CRLB is saturated with a maximum likelihood estimator under the correct model
+\item wrong spectrum template = wrong and noisy \RV s
+\end{itemize}
+
+\begin{itemize}
+\item what is the Cramer-Rao bound on a spectrum?
+\item Butler equation
+\item Bouchy equation
+\item Figueira (or Lovis \& Fischer) equation
+\end{itemize}
+
+In conclusion: the saturate the CRLB, we need to construct a very accurate spectral template, an accurate tellurics template, and maximize the likelihood.
+Given that theory sucks, this probably means data-driven.
+
 \section{Overview of Methods}
@@
 
-cross-correlation vs. max likelihood
+cross-correlation and max likelihood are the same: here's why. We are going to consider cross-correlation for simplicity in this paper.
 
@@
 \begin{itemize}
-\item ``binary'' mask
-\item synthetic spectrum (or template observation?)
-\item data-driven template spectrum
+\item ``binary'' mask: reasons you might want to use this (automatically ignores tellurics, trivially computes average line profile), pipelines that use this \& how (quote HARPS papers)
+\item synthetic spectrum (i.e.
+a slightly wrong model prediction, since the stellar model will never be a perfect reproduction of the star in question): sub-tweet: some surveys cross-correlate with a single (same) template for every single star, which is extremely wrong.
+you can be more right by using a spectral-type-specific template.
+\item data-driven template spectrum: some surveys (HIRES) use a single observation as a template. combining information from many visits is better because the combined SNR is amazing.
 \end{itemize}
 
-information theory based discussion of these: compute the CRLB for velocity. Mention how uncertainties are typically computed (Butler 1996 and Bouchy 2001 formulae).
+discuss quadratic max procedure to get a fine-grained maximum from a coarse-grained cross-correlation grid
 
@@
 
-\section{Testing with Toy-Model Spectra}
+\section{Data}
+
+Most of what we do in this paper uses artificial data because we need access to the truth in order to compute the theoretical information content of the data.
+As a final test, we do use real data as a demonstration that the methods presented here are useful when applied to real data and the assumptions made in our artificial data generation are realistic.
+
+\subsection{Artificial Data}
 
@@
 
-We repeated this synthesis to make a set of 512 spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion).
+We repeated this synthesis to make a set of \todo{64} spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion).
 The \RV s can be applied as a straightforward Doppler shift in the centers of each line.
@@
 
-\todo{remember to discuss quadratic max procedure.. probably in previous section?}
+discuss single-line and multi-line cases
 
-\subsection{Single-Line Results}
+FIGURE showing multi-line artificial data
+
+\subsection{HARPS Data}
+
+munging notes
+
+FIGURE showing analogous section of real data
+
+\section{Experiments}
+
+\subsection{Single-Line}
 
@@
 
-\subsection{Multi-Line Results}
+\subsection{Binary Mask}
 
@@
 
-\section{Dependence of RV Precision on Template Choice}
+\subsection{Rigid Template}
 
-how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template.
+how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template. do a correct template and an incorrect template.
 
-is this where we bring in the data-driven template and demonstrate how well it performs?
+even a tiny wrongness to the template is visible in the RV variance.
 
-\section{Correlated Noise Sources and Their Consequences}
+\subsection{Data-Driven Template}
+
+how do we construct the template? show that it comes close to saturating the CRLB.
+
+comment: there are far better ways to derive the data-driven template. we leave this to future work.
+
+\subsection{Realistic Sources of Structured Noise}
 
@@
 \begin{itemize}
-\item (micro-) telluric features
-\item subtle continuum trends
-\item cosmic rays?
+\item (micro-) telluric features: tellurics that you can't identify are important
+\item subtle continuum trends: not that big a deal
+\item cosmic rays?: we might skip this.
 \end{itemize}
 
-\section{Application to Real Spectra}
+\subsection{HARPS Data}
 
-how do our results compare to the reality of the situation? (will this tell us anything other than ``real spectra are more complicated and our template is silly''?)
+all the data-driven code works on real data.
+we get a better RMS on the star from the data-driven template than from the binary mask.
+based on this, we believe that the HARPS pipeline doesn't strictly use a binary mask.
 
-%\section{Discussion}
+\section{Summary \& Discussion}
 
-\section{Conclusions}
+paragraph for each major result from the abstract
+
+wrongness of our assumptions - how does this affect our conclusions?
 
 \acknowledgements
-MPIA hospitality
+MPIA hospitality, Montet, Foreman-Mackey, Bean, Sturmer, Rix, HARPS team, ESO, Stars group meeting, Brewer
 

### 2017-11-29 paper/paper.tex
additive

@@
 \item subtle continuum trends: not that big a deal
-\item cosmic rays?: we might skip this.
+\item cosmic rays?: we might skip this and add to discussion
+\item additive signals?: for example sky emission, lunar and other reflected solar; we might skip this and just add to discussion
 \end{itemize}
@@
 
+limitations of our ``additional noise'' ideas and what we have considered.
+
+Hogg: insert some grand final statement.
+
 \acknowledgements

### 2017-12-01 paper/paper.tex
crlb text

@@
 \newcommand{\CRLB}{\acronym{CRLB}}
+\newcommand{\EPRV}{\acronym{EPRV}}
 
@@
 
+Given finite, noisy data, there are limits coming from information
+theory on how well any parameter of interest can be measured.
+The most straightforward bound is the Cram\'er--Rao Lower Bound (\CRLB) on the
+variance of any unbiased frequentist estimator.
+Estimators can only beat this bound by taking on bias.
+
+The \CRLB\ is related to the Fisher Information that is frequently
+used in cosmological experiment forecasts.
+
+This bound is strong, but it cannot be calculated \emph{objectively};
+any description of the information in a data set requires
+subjective decisions (that is, debatable decisions), which are inputs
+to a model for the generation of the data.
+In particular, information-theory bounds require a model for the
+probability of the data given the parameter of interest; that is, they
+require a likelihood function.
+(Side note: Bayesians need a likelihood function for inference, and
+frequentists need a likelihood function for analyzing their estimators.
+The likelihood function is central to all statistical methods.)
+
+In the case of \EPRV, this says that it is impossible to compute the
+CRLB without making assumptions about the stellar spectrum (the
+latent, unobserved, \emph{true} spectrum that could only be observed
+with arbitrarily good data) and the noise processes that distort the
+spectrum in any finite, noisy observation.
+
+These noise processes necessarily must include any beliefs about
+unsubtracted telluric absorptions or emissions, and time variability
+of the star itself, not to mention photon and spectrograph read and
+extraction noise.
+
+The issue with
+
 \begin{itemize}
@@
 \begin{itemize}
-\item what is the Cramer-Rao bound on a spectrum?
+\item what is the Cramer-Rao bound on a spectrum? See notes in Hogg's red book for 2017-12-01.
 \item Butler equation

### 2017-12-01 paper/paper.tex
CRLB text

@@
 
-The \CRLB\ is related to the Fisher Information that is frequently
-used in cosmological experiment forecasts.
-
 This bound is strong, but it cannot be calculated \emph{objectively};
@@
 
-In the case of \EPRV, this says that it is impossible to compute the
-CRLB without making assumptions about the stellar spectrum (the
+In particular, the \CRLB\ is related to the Fisher Information that is
+frequently used in cosmological experiment forecasts (\todo{CITE
+  SHIT}).
+It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
+estimator must be larger than the second derivative of the
+\emph{negative} log likelihood function.
+In equations:
+\begin{eqnarray}
+\frac{1}{\sigma_v^2} &\geq& \frac{d^2}{d_v^2}(-\ln L)
+\end{eqnarray}
+
+In the case of \EPRV, this says that any computation of the limits
+on estimating stellar velocities (or stellar velocity changes)
+require making assumptions about the stellar spectrum (the
 latent, unobserved, \emph{true} spectrum that could only be observed
@@
 spectrum in any finite, noisy observation.
-
-These noise processes necessarily must include any beliefs about
+These noise processes ought (in principle) to include any beliefs about
 unsubtracted telluric absorptions or emissions, and time variability
@@
 extraction noise.
+However---and in keeping with the literature---we can make maximal or
+best-case assumptions, and ask what we would get for \EPRV\ precision
+in that best case.
+Then we can ask (with fake data) how much different choices about our
+data analysis, and different adversarial injections of noise,
+tellurics, spectrograph issues, and other wrongness can prevent us
+from achieving the best-case bounds.
 
-The issue with
+The most optimistic assumptions are... Gaussian... No
+tellurics... perfect calibration... no time dependence... all
+variances known...
 

### 2017-12-01 paper/paper.tex
intro edits

--- 2017-11-29 paper/paper.tex
@@
 
-While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of template
-
-We begin in Section \ref{s:methods} with a summary of the practices used by current \RV\ pipelines, including cross-correlation and template fitting.
-We then implement these techniques to extract \RV s from a 5-\r{A} section of simulated spectrum and compare their performance in Section \ref{s:fakedata}.
+While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of reference spectrum or template to the treatment of telluric absorption features.
+Ultra-stabilized spectrographs like HARPS \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
+...).
 
@@
 
-The assumption of this project is that we have many observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
+We begin in Section \ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum.
+In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting.
+We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section {s:data}.
+We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section {s:summary}.
+
+It is important to state at the outset the assumptions under which are operating for these experiments.
+We assume that our data set consists of multiple observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
 This means we are not considering the case of a real-time data reduction pipeline.
@@
 In other work, we will discuss this, but our current belief is that current RV spectrographs are supremely well wavelength-calibrated.
+
+[any other assumptions?]
 
@@
 \section{Data}
+\label{s:data}
 
@@
 \section{Experiments}
+\label{s:experiments}
 
@@
 \section{Summary \& Discussion}
+\section{s:summary}
 
@@
 \acknowledgements
-MPIA hospitality, Montet, Foreman-Mackey, Bean, Sturmer, Rix, HARPS team, ESO, Stars group meeting, Brewer
+MPIA hospitality, Montet, Foreman-Mackey, Bean, Stuermer, Rix, HARPS team, ESO, Stars group meeting, Brewer
 

### 2017-12-01 paper/paper.tex
math typo

--- 2017-12-01 paper/paper.tex
@@
 \newcommand{\ms}{\m\,\s^{-1}}
-\newcommand\todo[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
+\newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 \newcommand{\EPRV}{\acronym{EPRV}}
+\newcommand{\dd}{\mathrm{d}}
 
@@
 It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
-estimator must be larger than the second derivative of the
-\emph{negative} log likelihood function.
-In equations:
+estimator must be smaller than the second derivative of the
+\emph{negative} log likelihood function:
 \begin{eqnarray}
-\frac{1}{\sigma_v^2} &\geq& \frac{d^2}{d_v^2}(-\ln L)
+\frac{1}{\sigma_v^2} &\geq& \frac{\dd^2}{\dd v^2}(-\ln L)
 \end{eqnarray}

### 2017-12-01 paper/paper.tex
more background in intro

--- 2017-12-01 paper/paper.tex
@@
 Ultra-stabilized spectrographs like HARPS \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
-...).
+Queloz1995, Baranne1996).
+A lesser-used alternative is cross-correlation or maximum-likelihood fitting using a template spectrum generated from the data (Buchhave, Anglada-Escude).
+This often involves masking or downweighting of the most telluric-contaminated spectral regions.
 
-[summary of the ways people derive precise \RV s from spectra; citations to binary mask papers, papers that compare template matching to binary masks]
+On the other hand, gas-cell-calibrated spectrographs like HIRES and [whatever the McDonald one is] commonly adopt a more comprehensive forward-modeling approach.
+In this framework, the instrumental line spread function, the stellar RV, and other parameters are optimized by modeling each observation as a combination of shifted and instrument-convolved template spectra for the star and gas cell.
+[say a bit more]
+
+Our primary aim in this work is to establish a common frame of reference for these methods based on their relative ability to extract maximum-precision RVs and their robustness in the face of various expected structured noise sources.
 
@@
 In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting.
-We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section {s:data}.
-We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section {s:summary}.
+We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section \ref{s:data}.
+We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
 
@@
 \section{Summary \& Discussion}
-\section{s:summary}
+\label{s:summary}
 

### 2017-12-02 paper/paper.tex
keywords

@@
 
+\shorttitle{achieving maximum radial-velocity precision}
+\shortauthors{bedell and hogg}
+
 \begin{document}
@@
 \affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affiliation{University of Chicago, NEED DETAILS HERE}
+\affiliation{University of Chicago, \todo{need details here, and ORCID above}}
 
@@
 \end{abstract}
+
+\keywords{
+atmospheric effects
+---
+binaries: spectroscopic
+---
+methods: data analysis
+---
+methods: statistical
+---
+techniques: radial velocities
+---
+planets and satellites: detection
+}
 

### 2017-12-02 paper/paper.tex
cosmetics
more on CRLB

@@
 
+% units macros
 \newcommand{\unit}[1]{\mathrm{#1}}
@@
 \newcommand{\ms}{\m\,\s^{-1}}
+
+% math macros
+\newcommand{\dd}{\mathrm{d}}
+\newcommand{\T}{^{\mathsf{T}}}
+
+% text macros
 \newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
@@
 \newcommand{\EPRV}{\acronym{EPRV}}
-\newcommand{\dd}{\mathrm{d}}
 
+\setlength{\parindent}{1.4em}
 \shorttitle{achieving maximum radial-velocity precision}
@@
 
-\begin{abstract}
+\begin{abstract}\noindent
 % context
@@
 
-Given finite, noisy data, there are limits coming from information
+Given finite, noisy data, there is a limit coming from information
 theory on how well any parameter of interest can be measured.
-The most straightforward bound is the Cram\'er--Rao Lower Bound (\CRLB) on the
-variance of any unbiased frequentist estimator.
+This limit is
+sometimes known as the Cram\'er--Rao Lower Bound (\CRLB; \todo{cite CR and wikipedia}).
+In its simplest form, it constrains the variance of any unbiased
+frequentist estimator.
 Estimators can only beat this bound by taking on bias.
@@
 require a likelihood function.
-(Side note: Bayesians need a likelihood function for inference, and
+As an aside: Bayesians need a likelihood function for inference, and
 frequentists need a likelihood function for analyzing their estimators.
-The likelihood function is central to all statistical methods.)
+The likelihood function is therefore central to all statistical methods,
+and the core component of any data analysis.
 
@@
 It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
-estimator must be smaller than the second derivative of the
-\emph{negative} log likelihood function:
-\begin{eqnarray}
-\frac{1}{\sigma_v^2} &\geq& \frac{\dd^2}{\dd v^2}(-\ln L)
-\end{eqnarray}
+estimator (that is, the information in the velocity) must be smaller than the
+second derivative of the \emph{negative} log likelihood function:
+\begin{equation}
+\frac{1}{\sigma_v^2} \leq E\left[\frac{\dd^2}{\dd v^2}(-\ln L)\right] \quad,
+\end{equation}
+where $L$ is the likelihood, or the probability of the data given the velocity,
+and the expectation $E[]$ is taken over all possible data (that is, it is an
+expectation under the likelihood).
+If there are many other nuisance parameters, a Bayesian can marginalize them
+out to make $L$ the marginalized likelihood for the velocity, or a frequentist
+can include them in the estimation and compute an inverse covariance matrix $C^{-1}$
+for the full parameter vector, which will have the bound
+\begin{equation}
+v^T\cdot C^{-1}\cdot v \leq v^T\cdot Q\cdot v
+\end{equation}
+\begin{equation}
+Q_{ij} \equiv E\left[\frac{\dd^2}{\dd\theta_i\,\dd\theta_j}(-\ln L)\right] \quad,
+\end{equation}
+where $Q$ is the information tensor, $i,j$ are indices into that
+tensor, corresponding parameters are $\theta_i, \theta_j$, and $v$ is
+any arbitrary vector.  That is, the bound is not just on every
+parameter, but really on on any combination of parameters.
 
@@
 on estimating stellar velocities (or stellar velocity changes)
-require making assumptions about the stellar spectrum (the
+requires a likelihood function.
+This, in turn, requires making assumptions about the stellar spectrum (the
 latent, unobserved, \emph{true} spectrum that could only be observed
@@
 \begin{itemize}
-\item the Cramer-Rao bound is the limit on measurement
-\item it relates to Fisher information
-\item it cannot be computed without a model
+\item what is the Cramer-Rao bound on a spectrum in optimistic case? 
 \item CRLB is saturated with a maximum likelihood estimator under the correct model
 \item wrong spectrum template = wrong and noisy \RV s
-\end{itemize}
-
-\begin{itemize}
-\item what is the Cramer-Rao bound on a spectrum? See notes in Hogg's red book for 2017-12-01.
-\item Butler equation
+\item Butler equation \todo{See notes in Hogg's red book for 2017-12-01.}
 \item Bouchy equation
-\item Figueira (or Lovis \& Fischer) equation
+\item Figueira (or Lovis \& Fischer)
 \end{itemize}

### 2017-12-02 paper/paper.tex
fixing typos; generalizing CRLB

@@
 It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
-estimator (that is, the information in the velocity) must be smaller than the
+estimator (that is, the information in the velocity estimator) must be smaller than the
 second derivative of the \emph{negative} log likelihood function:
@@
 where $L$ is the likelihood, or the probability of the data given the velocity,
-and the expectation $E[]$ is taken over all possible data (that is, it is an
+and the expectation $E[\cdot]$ is taken over all possible data (that is, it is an
 expectation under the likelihood).

### 2017-12-03 paper/paper.tex
optimistic CRLB

@@
 
-The most optimistic assumptions are... Gaussian... No
-tellurics... perfect calibration... no time dependence... all
-variances known...
+The most optimistic assumptions we can possibly make are the following.
+\begin{itemize}
+\item The spectrograph is perfectly calibrated in a wavelength sense.
+\item The spectrograph is consistently and repeatably calibrated in a
+  flux sense. That is, the spectrum does not need to be perfectly
+  continuum-normalized or perfectly flux-normalized, but it is
+  calibrated near-perfectly consistently across epochs (exposures).
+\item There are no residual telluric absorptions nor any residual sky
+  emission in the spectra. That is, the sky is near-perfectly calibrated
+  and removed.
+\item There is no time dependence (no epoch-to-epoch variations) in the
+  stellar spectrum.
+\item The \emph{true} spectrum (that is, the latent spectrum that
+  would be observed if the data were far, far better) of the star is
+  known at better accuracy and precision than the main noise sources.
+\item All noise sources contributing to the spectral data---which
+  include at a minimum photon noise, read noise, residual telluric
+  issues, residual wavelength-calibration issues, and residual stellar
+  spectral variability---are, when summed together, indistinguishable
+  from being zero-mean, Gaussian, and additive.
+\item The complete variance tensor of the total noise on the spectrum is known, and
+  known correctly. That is, the variance of the noise at every pixel
+  is known, as is the covariance between nearby pixels, and in the
+  same \emph{true} sense that the true spectrum of the star is known.
+\end{itemize}
+It is worthy of note that every formula (that we know) in the
+literature for maximum \RV\ precision (\todo{cite Butler, Bouchy,
+  Figueira, Lovis}) makes these same assumptions, plus additional \emph{even
+more restrictive assumptions}.
+Under these disturbingly unrealistic assumptions, the \CRLB\ becomes
+\begin{equation}
+\frac{1}{\sigma_v^2} \leq \left[\frac{\dd f}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f}{\dd v}\right]
+\end{equation}
+where the derivatives are of the true spectrum with respect to
+velocity (which can be thought of as the derivative of a Doppler
+operator with respect to velocity acting on the true spectrum), the
+derivatives are column vectors, and $C^{-1}$ is the covariance matrix
+describing the true variance and covariances of the noise process.
 
+...Now speak to the more restrictive assumptions others have made but
+say that we agree in the end.
 \begin{itemize}
-\item what is the Cramer-Rao bound on a spectrum in optimistic case? 
-\item CRLB is saturated with a maximum likelihood estimator under the correct model
-\item wrong spectrum template = wrong and noisy \RV s
 \item Butler equation \todo{See notes in Hogg's red book for 2017-12-01.}
@@
 
-In conclusion: the saturate the CRLB, we need to construct a very accurate spectral template, an accurate tellurics template, and maximize the likelihood.
-Given that theory sucks, this probably means data-driven.
+... Now note that the \CRLB\ is saturated with a maximum likelihood estimator under the correct model.
+
+... Now note that using a wrong spectrum template will necessarily move you away from the bound. Bad!
+
+In conclusion: the saturate the CRLB, we need to construct a very
+accurate spectral template, an accurate tellurics template, and
+maximize the likelihood. Given that theory sucks, this probably means
+data-driven.
 

### 2017-12-03 paper/paper.tex
typo

@@
 operator with respect to velocity acting on the true spectrum), the
-derivatives are column vectors, and $C^{-1}$ is the covariance matrix
+derivatives are column vectors, and $C^{-1}$ is the inverse of the covariance matrix
 describing the true variance and covariances of the noise process.

### 2017-12-04 paper/paper.tex
info section done - boo yeah - write drunk, edit sober

@@
 Under these disturbingly unrealistic assumptions, the \CRLB\ becomes
-\begin{equation}
+\begin{equation}\label{eq:crlb}
 \frac{1}{\sigma_v^2} \leq \left[\frac{\dd f}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f}{\dd v}\right]
@@
 
-...Now speak to the more restrictive assumptions others have made but
-say that we agree in the end.
-\begin{itemize}
-\item Butler equation \todo{See notes in Hogg's red book for 2017-12-01.}
-\item Bouchy equation
-\item Figueira (or Lovis \& Fischer)
-\end{itemize}
+Formulae (bounds) related to (\ref{eq:crlb}) have appeared previously
+in the literature, although in somewhat more complicated forms.
+Butler (\todo{cite}) shows a similar expression, but makes the additional
+assumption that every pixel is independent; that is, that there is no
+covariance between nearby pixels from, say, extraction or continuum normalization.
+Bouchy (\todo{cite}) shows an expression that assumes that the pixels
+are independent and that the noise is dominated by a combination of photon
+and read noise. 
+That work also replaces the model spectrum with a set of optimal
+weights for cross-correlation, which amounts to very much the same thing.
+Figueira (\todo{cite}) and \todo{others (cite)} show a per-line
+bound, based on assumptions similar to those of Bouchy but with the additional assumption that
+each line is well approximated by a Gaussian (squared exponential) in shape.
+These literature bounds are all extremely similar; nothing here is controversial.
 
-... Now note that the \CRLB\ is saturated with a maximum likelihood estimator under the correct model.
+Now that we have established the bound, how can we saturate it? That is,
+how can we make measurements (estimators)  with variances that come close to the \CRLB?
+The short answer is that reliable bound-saturating estimators are maximum-likelihood estimators.
+The long answer is long, but part of it is that the bound can only be saturated if the
+assumptions are correct, and note that the assumptions include that we know the
+\emph{true} spectrum of the star, among many other things.
+That is, if we can take data that come close to meeting the assumptions, and
+if we can find a spectral model that comes close to the true spectrum of the star,
+then in principle we can estimate radial velocities with a precision close to the \CRLB.
 
-... Now note that using a wrong spectrum template will necessarily move you away from the bound. Bad!
+Given that computational stellar spectral models have physical issues, the best way
+to get an accurate estimate of the true spectrum of the star is going to be data-driven.
+In what follows, we are going to show that---for a typical \HARPS-like observing program---%
+it is possible to build a data-driven model of the stellar spectrum that is good enough
+to nearly-saturate the \CRLB.
+That is, with a data-driven model of the true stellar spectrum, model inaccuracy will
+not dominate the \RV\ noise budget.
+We will additionally bring evidence that telluric features can be a significant problem,
+but that they also can plausibly be mitigated with a data-driven model.
 
-In conclusion: the saturate the CRLB, we need to construct a very
-accurate spectral template, an accurate tellurics template, and
-maximize the likelihood. Given that theory sucks, this probably means
-data-driven.
+One final comment:
+We are forced to perform these experiments primarily with artificial data.
+The reason is: In order to assess the performance with respect to latent, unobservable \emph{truth},
+we require knowledge of that latent, unobservable truth.
+When we make artificial data, we play the r\^ole of God.
+We will also show tests on two real stars, one hosting a companion and
+one (apparently) companion-free (at least at our precision).
 

### 2017-12-04 paper/paper.tex
tweak first sentences

@@
 
-Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from galactic dynamics to cosmic expansion.
-In recent decades, however, a new focus has emerged on \textit{extremely precise} \RV\ measurements of stars as a means of detecting exoplanets.
-For this application, measuring the absolute velocity of the star is much less important than resolving its changes through time.
+Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from Galactic dynamics to large-scale structure and cosmic expansion.
+Fundamentally, accuracy of radial-velocity mesurement is limited by the quality of spectral models of stars, or templates.
+In recent decades, however, a new focus has emerged on \emph{extremely precise} but purely \emph{relative} \RV\ measurements of stars as a means of detecting exoplanets and other kinds of companions.
+For this application, measuring the absolute velocity of the star is much less important than resolving changes through time.
 With sub-$\ms$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.

### 2017-12-04 paper/paper.tex
intro tweka

@@
 At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
-In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision \RV s: the method of analysis used to extract the \RV\ from a stellar spectrum.
+In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision \RV s: the method of analysis used to estimate the \RV\ given an extracted stellar spectrum.
 

### 2017-12-04 paper/paper.tex
edits

@@
 
-\setlength{\parindent}{1.4em}
+\setlength{\parindent}{1.4em} % trust in Hogg
 \shorttitle{achieving maximum radial-velocity precision}
@@
 
-\begin{document}
+\begin{document}\sloppy\sloppypar\raggedbottom\frenchspacing % trust in Hogg
 \graphicspath{ {figures/} }
@@
   continuum-normalized or perfectly flux-normalized, but it is
-  calibrated near-perfectly consistently across epochs (exposures).
+  calibrated near-perfectly \emph{consistently} across epochs (exposures).
 \item There are no residual telluric absorptions nor any residual sky
@@
 We will also show tests on two real stars, one hosting a companion and
-one (apparently) companion-free (at least at our precision).
+one (apparently) companion-free (at least at our precision), but these
+tests will necessarily only be confirmatory of the more stringent results
+from artificial data.
 

### 2017-12-04 paper/paper.tex
minor changes

--- 2017-12-04 paper/paper.tex
@@
 
-\author{Megan Bedell}
+\author[0000-0001-9907-7742]{Megan Bedell}
 \affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affiliation{University of Chicago, \todo{need details here, and ORCID above}}
+\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
 
@@
 
-Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from Galactic dynamics to large-scale structure and cosmic expansion.
-Fundamentally, accuracy of radial-velocity mesurement is limited by the quality of spectral models of stars, or templates.
-In recent decades, however, a new focus has emerged on \emph{extremely precise} but purely \emph{relative} \RV\ measurements of stars as a means of detecting exoplanets and other kinds of companions.
-For this application, measuring the absolute velocity of the star is much less important than resolving changes through time.
+Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from galactic dynamics to cosmic expansion. 
+In recent decades, however, a new focus has emerged on \textit{extremely precise} \RV\ measurements of stars as a means of detecting exoplanets. 
+For this application, measuring the absolute velocity of the star is much less important than resolving its changes through time. 
 With sub-$\ms$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
@@
 
-We begin in Section \ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum.
-In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting.
-We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section \ref{s:data}.
-We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
-
 It is important to state at the outset the assumptions under which are operating for these experiments. 
@@
 This means we are not considering the case of a real-time data reduction pipeline. 
-We also assume a HARPS-like setup, i.e.
-no absorption cell, but the points we make can be easily generalized to the gas-cell case.
+We also assume a HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. 
 We return to this point in the discussion \todo{(do this)}.
@@
 [any other assumptions?]
+
+We begin in Section \ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum. 
+In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting. 
+We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section \ref{s:data}. 
+We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
 
@@
 Under these disturbingly unrealistic assumptions, the \CRLB\ becomes
-\begin{equation}\label{eq:crlb}
+\begin{equation}
 \frac{1}{\sigma_v^2} \leq \left[\frac{\dd f}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f}{\dd v}\right]
@@
 
-Formulae (bounds) related to (\ref{eq:crlb}) have appeared previously
-in the literature, although in somewhat more complicated forms.
-Butler (\todo{cite}) shows a similar expression, but makes the additional
-assumption that every pixel is independent; that is, that there is no
-covariance between nearby pixels from, say, extraction or continuum normalization.
-Bouchy (\todo{cite}) shows an expression that assumes that the pixels
-are independent and that the noise is dominated by a combination of photon
-and read noise. 
-That work also replaces the model spectrum with a set of optimal
-weights for cross-correlation, which amounts to very much the same thing.
-Figueira (\todo{cite}) and \todo{others (cite)} show a per-line
-bound, based on assumptions similar to those of Bouchy but with the additional assumption that
-each line is well approximated by a Gaussian (squared exponential) in shape.
-These literature bounds are all extremely similar; nothing here is controversial.
+...Now speak to the more restrictive assumptions others have made but
+say that we agree in the end.
+\begin{itemize}
+\item Butler equation \todo{See notes in Hogg's red book for 2017-12-01.}
+\item Bouchy equation
+\item Figueira (or Lovis \& Fischer)
+\end{itemize}
 
-Now that we have established the bound, how can we saturate it? That is,
-how can we make measurements (estimators)  with variances that come close to the \CRLB?
-The short answer is that reliable bound-saturating estimators are maximum-likelihood estimators.
-The long answer is long, but part of it is that the bound can only be saturated if the
-assumptions are correct, and note that the assumptions include that we know the
-\emph{true} spectrum of the star, among many other things.
-That is, if we can take data that come close to meeting the assumptions, and
-if we can find a spectral model that comes close to the true spectrum of the star,
-then in principle we can estimate radial velocities with a precision close to the \CRLB.
+... Now note that the \CRLB\ is saturated with a maximum likelihood estimator under the correct model.
 
-Given that computational stellar spectral models have physical issues, the best way
-to get an accurate estimate of the true spectrum of the star is going to be data-driven.
-In what follows, we are going to show that---for a typical \HARPS-like observing program---%
-it is possible to build a data-driven model of the stellar spectrum that is good enough
-to nearly-saturate the \CRLB.
-That is, with a data-driven model of the true stellar spectrum, model inaccuracy will
-not dominate the \RV\ noise budget.
-We will additionally bring evidence that telluric features can be a significant problem,
-but that they also can plausibly be mitigated with a data-driven model.
+... Now note that using a wrong spectrum template will necessarily move you away from the bound. Bad!
 
-One final comment:
-We are forced to perform these experiments primarily with artificial data.
-The reason is: In order to assess the performance with respect to latent, unobservable \emph{truth},
-we require knowledge of that latent, unobservable truth.
-When we make artificial data, we play the r\^ole of God.
-We will also show tests on two real stars, one hosting a companion and
-one (apparently) companion-free (at least at our precision).
+In conclusion: the saturate the CRLB, we need to construct a very
+accurate spectral template, an accurate tellurics template, and
+maximize the likelihood. Given that theory sucks, this probably means
+data-driven.
 

### 2017-12-04 paper/paper.tex
writing data section

--- 2017-12-04 paper/paper.tex
@@
 
-One final comment:
-We are forced to perform these experiments primarily with artificial data.
-The reason is: In order to assess the performance with respect to latent, unobservable \emph{truth},
-we require knowledge of that latent, unobservable truth.
-When we make artificial data, we play the r\^ole of God.
-We will also show tests on two real stars, one hosting a companion and
-one (apparently) companion-free (at least at our precision), but these
-tests will necessarily only be confirmatory of the more stringent results
-from artificial data.
-
 \section{Overview of Methods}
@@
 
-Most of what we do in this paper uses artificial data because we need access to the truth in order to compute the theoretical information content of the data.
-As a final test, we do use real data as a demonstration that the methods presented here are useful when applied to real data and the assumptions made in our artificial data generation are realistic.
+The experiments we perform in this paper primarily use artificial data, for the simple reason that we require access to the underlying true RVs in order to assess the performance of our methods. 
+We attempt to generate the fake data as realistically as possible for the case of a high-resolution, high-signal-to-noise spectrum of a Sun-like star taken with \HARPS. 
+With these data, we are able to compute the \CRLB\. 
+We can also isolate the effects of various confounding noise sources like telluric features and imperfect continuum normalization by injecting them into the artificial data, which is generated free of correlated noise by default.
+
+We will also show tests on real \HARPS\ data for two stars, one hosting a companion and
+one (apparently) companion-free (at least at our precision). 
+These tests serve as a demonstration that the methods presented here are useful when applied to real data and the assumptions made in our artificial data generation are realistic.
 
@@
 We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single \RV\ spectrum. 
-Lines were inserted as Gaussians with arbitrarily-prescribed centers and depths (or equivalent widths).
+Lines were inserted as Gaussians with prescribed centers, widths, and depths. 
 Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 $\AA$. 
-%The number of lines used varies in the results discussed below.
+In the simplest test case, an arbitrary single line at 5000 $\AA$ was used (see Section \ref{s:singleline}). 
+However, for most of our tests we injected a set of 12 lines with centers between 4998 - 5002 $\AA$. 
+The Gaussian parameters for these lines were based on fits to a reflected Solar spectrum from Vesta taken by \HARPS. \todo{(find citation/program number)} 
+The resulting spectrum strongly resembles a 5-$\AA$ chunk taken from a typical \HARPS\ spectrum of a Sun-like star (Figure \ref{fig:spectra}).
 
 We repeated this synthesis to make a set of \todo{64} spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion).
-The \RV s can be applied as a straightforward Doppler shift in the centers of each line.
-\todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.
-
-discuss single-line and multi-line cases
-
-FIGURE showing multi-line artificial data
+The \RV s were applied as a multiplicative Doppler shift on the central wavelength of each spectral line to be injected. 
+At this point, we are able to run a cross-correlation analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.
 
@@
 
-FIGURE showing analogous section of real data
+reference to figure comparing real \& fake data
 
@@
 \subsection{Single-Line}
+\label{s:singleline}
 

### 2017-12-04 paper/paper.tex
minor edits

@@
 \newcommand{\ms}{\m\,\s^{-1}}
+\newcommand{\ang}{\text{\normalfont\AA}}
 
@@
 While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of reference spectrum or template to the treatment of telluric absorption features.
-Ultra-stabilized spectrographs like HARPS \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
+Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
 Queloz1995, Baranne1996).
@@
 
-It is important to state at the outset the assumptions under which are operating for these experiments. 
-We assume that our data set consists of multiple observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
+For the purposes of this work, we assume that our data set consists of multiple observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
 This means we are not considering the case of a real-time data reduction pipeline. 
@@
 
-We additionally assume that the data are extremely well wavelength-calibrated. 
-In other work, we will discuss this, but our current belief is that current RV spectrographs are supremely well wavelength-calibrated.
-
-[any other assumptions?]
+%wavelength calibration assumption removed b/c it fits better in Section 2... but maybe we want to add it back?
 
@@
 In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting. 
-We then implement these techniques to extract \RV s using 5-\r{A} section of simulated and real spectra, as described in Section \ref{s:data}. 
+We then implement these techniques to extract \RV s using 5-\ang-wide sections of simulated and real spectra, as described in Section \ref{s:data}. 
 We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
@@
 The basic model that we use to generate fake spectra is extremely simple, assuming a perfectly normalized continuum; isolated, perfectly Gaussian spectral lines; and white noise only. 
-We simulated a 5-$\AA$ region with random noise corresponding to an SNR of 100, a typical value for a single \RV\ spectrum. 
+We simulated a 5-\ang\ region with random noise corresponding to an SNR of 100, a typical value for a single \RV\ spectrum. 
 Lines were inserted as Gaussians with prescribed centers, widths, and depths. 
-Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 $\AA$. 
-In the simplest test case, an arbitrary single line at 5000 $\AA$ was used (see Section \ref{s:singleline}). 
-However, for most of our tests we injected a set of 12 lines with centers between 4998 - 5002 $\AA$. 
+Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 \ang. 
+In the simplest test case, an arbitrary single line at 5000 \ang\ was used (see Section \ref{s:singleline}). 
+However, for most of our tests we injected a set of 12 lines with centers between 4998 - 5002 \ang. 
 The Gaussian parameters for these lines were based on fits to a reflected Solar spectrum from Vesta taken by \HARPS. \todo{(find citation/program number)} 
-The resulting spectrum strongly resembles a 5-$\AA$ chunk taken from a typical \HARPS\ spectrum of a Sun-like star (Figure \ref{fig:spectra}).
+The resulting spectrum strongly resembles a 5-\ang\ chunk taken from a typical \HARPS\ spectrum of a Sun-like star (Figure \ref{fig:spectra}).
 
@@
 
-\subsection{HARPS Data}
+\begin{figure}
+\centering
+% \includegraphics[width=\columnwidth]{spectra}
+\caption{}
+\label{fig:spectra}
+\end{figure}
+
+\subsection{\HARPS\ Data}
 
@@
 
-\subsection{HARPS Data}
+\subsection{Real Data}
 
@@
 we get a better RMS on the star from the data-driven template than from the binary mask.
-based on this, we believe that the HARPS pipeline doesn't strictly use a binary mask.
+based on this, we believe that the \HARPS\ pipeline doesn't strictly use a binary mask.
 
@@
 \acknowledgements
-MPIA hospitality, Montet, Foreman-Mackey, Bean, Stuermer, Rix, HARPS team, ESO, Stars group meeting, Brewer
+MPIA hospitality, Montet, Foreman-Mackey, Bean, Stuermer, Rix, \HARPS\ team, ESO, Stars group meeting, Brewer
 

### 2017-12-08 paper/paper.tex
tweak
warning

@@
 each line is well approximated by a Gaussian (squared exponential) in shape.
-These literature bounds are all extremely similar; nothing here is controversial.
+These literature bounds are all extremely similar, and similar to (\ref{eq:crlb});
+nothing here is controversial.
 
@@
 Our next test involved a more realistic test spectrum.
-We measured the centers and depths of \todo{N} lines in the Solar spectrum between {x - y} $\AA$ and used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
+We measured the centers and depths of \todo{N} lines in the Solar spectrum between {x - y} $\ang$ and used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
 

### 2017-12-08 paper/paper.tex
x-corr == ML

@@
 
-cross-correlation and max likelihood are the same: here's why. We are going to consider cross-correlation for simplicity in this paper.
+Most methods for radial-velocity estimation involve some kind of
+cross-correlation of the data with a template.
+This might be surprising, since (as we noted above) bound-saturating
+estimators will be maximum-likelihood estimators.
+However, under some simple assumptions, cross-correlation optima are
+also maximum-likelihood optima.
 
-for either of these, you could use:
+For example, when the likelihood is Gaussian, the log-likelihood
+contains a chi-squared-like term
+\begin{equation}
+-\frac{1}{2}\,[y - f]\T\cdot C^{-1}\cdot [y - f] \quad ,
+\end{equation}
+where $y$ is a column vector of the data, $f$ is the model spectrum,
+and $C$ is the covariance matrix describing the noise variance.
+In the case of independent pixel noise this just becomes an
+inverse-variance weighted sum of squared residuals (chi-squared).
+If we expand this quadratic form, it becomes
+\begin{equation}
+-\frac{1}{2}\,[y - f]\T\cdot C^{-1}\cdot [y - f] = -\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f 
+                                                   -\frac{1}{2}\,f\T\cdot C^{-1}\cdot f \quad.
+\end{equation}
+That is, the chi-squared term factors into a term that is the square
+of the model, a term that is the square of the data, and a term that
+is a cross-correlation.
+As long as the amplitude of $f$ doesn't depend on velocity, and the
+$C$ tensor is ``smooth'' in certain respects, the dependence of this
+chi-squared term on velocity will be dominated by the
+cross-correlation.
+For this reason, we will use a variant of cross-correlation
+optimization as our radial-velocity estimator.
+
+...In detail, the cross-correlation we will optimize is...
+
+...For the cross-correlation template $f$ you could use...
 \begin{itemize}
@@
 
-discuss quadratic max procedure to get a fine-grained maximum from a coarse-grained cross-correlation grid
+....discuss quadratic max procedure to get a fine-grained maximum from a coarse-grained cross-correlation grid
 

### 2017-12-08 paper/paper.tex
typesetting
more text about template choice

@@
 \newcommand{\HARPS}{\project{\acronym{HARPS}}}
+\newcommand{\HIRES}{\project{\acronym{HIRES}}}
 \newcommand{\RV}{\acronym{RV}}
@@
 
-On the other hand, gas-cell-calibrated spectrographs like HIRES and [whatever the McDonald one is] commonly adopt a more comprehensive forward-modeling approach. 
+On the other hand, gas-cell-calibrated spectrographs like \HIRES\ commonly adopt a more comprehensive forward-modeling approach. 
 In this framework, the instrumental line spread function, the stellar RV, and other parameters are optimized by modeling each observation as a combination of shifted and instrument-convolved template spectra for the star and gas cell.
@@
 This means we are not considering the case of a real-time data reduction pipeline. 
-We also assume a HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. 
+We also assume a \HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. 
 We return to this point in the discussion \todo{(do this)}.
@@
 \begin{equation}
--\frac{1}{2}\,[y - f]\T\cdot C^{-1}\cdot [y - f] = -\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f 
+{}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f 
 -\frac{1}{2}\,f\T\cdot C^{-1}\cdot f \quad.
@@
 
-...For the cross-correlation template $f$ you could use...
+A variety of choices exist for the cross-correlation template $f$. A few possibilities include:
 \begin{itemize}
-\item ``binary'' mask: reasons you might want to use this (automatically ignores tellurics, trivially computes average line profile), pipelines that use this \& how (quote HARPS papers)
+\item a ``binary'' mask, or a spectrum of zero value with non-zero peaks at the expected rest-frame wavelengths of the spectral lines of interest.
+A cross-correlation between this mask and the observed spectrum will effectively compute the average line profile with its minimum corresponding to the best-fit RV.
+This method has the benefit of trivially excluding telluric-contaminated or otherwise unreliable regions of the spectrum by deleting them from the mask.
+A desired relative weighting of lines can be incorporated by setting the non-zero mask values to different amplitudes at different lines (cite Pepe).
+This method is currently used by the \HARPS\ pipeline (cite).
 \item synthetic spectrum (i.e.
@@
 We attempt to generate the fake data as realistically as possible for the case of a high-resolution, high-signal-to-noise spectrum of a Sun-like star taken with \HARPS. 
-With these data, we are able to compute the \CRLB\. 
+With these data, we are able to compute the \CRLB.  
 We can also isolate the effects of various confounding noise sources like telluric features and imperfect continuum normalization by injecting them into the artificial data, which is generated free of correlated noise by default.

### 2017-12-10 paper/paper.tex
affil
quadratic interpolation

@@
 \affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affiliation{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
+\affil{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
 
 \author[0000-0003-2866-9403]{David W. Hogg}
+\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
@@
 We return to this point in the discussion \todo{(do this)}.
-
-%wavelength calibration assumption removed b/c it fits better in Section 2... but maybe we want to add it back?
 
@@
 
-....discuss quadratic max procedure to get a fine-grained maximum from a coarse-grained cross-correlation grid
+One final implementation detail is the following:
+When we optimize the cross-correlation function, we don't run an optimizer;
+we perform brute-force search on a linear velocity grid.
+This grid, however, does not need to be fine.
+We use a coarse grid (\todo{give grid spacing here}), which does not
+over-resolve the velocity peak at the optimum.
+We then take the highest value on the grid and the values on either
+side, fit a parabola through those three values, and take as the optimal
+velocity the peak of that parabola.
+That is, we obtain sub-grid velocity resolution by means of a quadratic
+interpolation at the peak on the finite grid.
+\todo{Do we need to demonstrate this with a toy figure?}
 

### 2017-12-10 paper/paper.tex
cosmetics
what happens if the spectrum varies - Cthulhu

@@
 \newcommand{\CRLB}{\acronym{CRLB}}
-\newcommand{\EPRV}{\acronym{EPRV}}
 
@@
 
-In the case of \EPRV, this says that any computation of the limits
+In the case of extreme-precision \RV, this says that any computation of the limits
 on estimating stellar velocities (or stellar velocity changes)
@@
 However---and in keeping with the literature---we can make maximal or
-best-case assumptions, and ask what we would get for \EPRV\ precision
+best-case assumptions, and ask what we would get for \RV\ precision
 in that best case.
@@
 
-\section{Overview of Methods}
+\section{Radial-velocity estimators}
 \label{s:methods}
@@
 
-\section{Experiments}
+\section{Experiments and results}
 \label{s:experiments}
@@
 
-\section{Summary \& Discussion}
+\section{Summary and discussion}
 \label{s:summary}
 
-paragraph for each major result from the abstract
+...paragraph for each major result from the abstract
 
-wrongness of our assumptions - how does this affect our conclusions?
+In \sectionname~\ref{s:information} we listed some exceedingly
+optimistic assumptions under which we could compute the
+information-theoretic limit on radial-velocity precision.
+Some of these assumptions are worthy of discussion: How true are they,
+and what happens, both from an information-theory perspective and from
+a methodological perspective, when we relax them?
 
-limitations of our ``additional noise'' ideas and what we have considered.
+One of these assumptions, for example, is that the true stellar
+spectrum is constant with time; that is, the variations from
+epoch-to-epoch are from noise or nuisances (like telluric absorption).
+This assumption is critical and deep.
+The entire edifice of extreme-precision \RV\ is built on this assumption:
+With imperfect models of stellar structure and photospheres, it is
+impossible to know the \emph{absolute} \RV\ of a star to high precision.
+The reason is that small issues with (say) convective line shifts or
+line-shapes, it is impossible to know precisely the true wavelength of
+a line centroid, unless the model is correct in an absolute sense.
+So we satisfy ourselves that we can measure \emph{relative} velocities
+to arbitrarily high precision:
+Relative precision---and everything in this paper---depends on our
+ability to measure shifts from epoch to epoch.
+If the stellar spectrum is permitted (or observed) to change from epoch
+to epoch, there is no fundamental way to determine or confirm spectral
+shifts corresponding to Doppler shifts.
+That is, even relative \RV\ measurements depend on a constancy of the
+stellar spectrum.
+This is a deep issue, extending far beyond the scope of this paper;
+time-variable stellar spectra are not killing to the extreme-precision
+\RV\ program!
+But they do require substantial and non-trivial adjustments to the
+\foreign{status quo}.
+We hope to address these in future work.
 
-Hogg: insert some grand final statement.
+...MOAR assumptions
+
+...limitations of our ``additional noise'' ideas and what we have considered.
+
+...Hogg: insert some grand final statement.
 

### 2017-12-10 paper/paper.tex
bugfixes

@@
 % text macros
+\newcommand{\sectionname}{Section}
 \newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
@@
 \newcommand{\project}[1]{\textsl{#1}}
+\newcommand{\foreign}[1]{\textsl{#1}}
 \newcommand{\HARPS}{\project{\acronym{HARPS}}}
@@
 
-In \sectionname~\ref{s:information} we listed some exceedingly
+In \sectionname~\ref{s:info} we listed some exceedingly
 optimistic assumptions under which we could compute the

### 2017-12-11 paper/paper.tex
adjust wording

@@
 With imperfect models of stellar structure and photospheres, it is
-impossible to know the \emph{absolute} \RV\ of a star to high precision.
-The reason is that small issues with (say) convective line shifts or
-line-shapes, it is impossible to know precisely the true wavelength of
-a line centroid, unless the model is correct in an absolute sense.
-So we satisfy ourselves that we can measure \emph{relative} velocities
-to arbitrarily high precision:
+impossible to know the \emph{absolute} \RV\ of a star to high precision;
+we satisfy ourselves that we can measure \emph{relative} velocities
+to arbitrarily high precision.
 Relative precision---and everything in this paper---depends on our
@@
 stellar spectrum.
-This is a deep issue, extending far beyond the scope of this paper;
+This is a deep issue, extending far beyond the scope of this paper.
 time-variable stellar spectra are not killing to the extreme-precision
 \RV\ program!
-But they do require substantial and non-trivial adjustments to the
+But at current practices they may be limiting precision,
+and to ameliorate their effects will require substantial and non-trivial adjustments to the
 \foreign{status quo}.

### 2017-12-11 paper/paper.tex
typo
assumptions about the noise

@@
+% This file is part of the EPRV project.
+% Copyright 2017, 2018 the authors.
+
+% Style notes
+% - Re-define acronyms in each section? So less flippy flippy?
+
 \documentclass[modern]{aastex61}
@@
 This is a deep issue, extending far beyond the scope of this paper.
-time-variable stellar spectra are not killing to the extreme-precision
+Time-variable stellar spectra are not killing to the extreme-precision
 \RV\ program!
@@
 We hope to address these in future work.
+
+We also assumed that the noise is zero-mean, Gaussian, and with known
+variance tensor.
+There are certainly sources of noise that might violate these
+assumptions.
+There are two ways in which we might weaken our assumptions about
+the noise.
+The first is that the total noise acting on the system---and recall
+that noise for us is anything not accounted for in our model of the
+data---might not be Gaussian in form.
+This will happen if the photon counts (per pixel) get low.
+But it can also happen if there are unmodeled calibration artifacts
+that don't happen to have Gaussian-distributed amplitudes.
+This is not a problem in principle with computing or saturating the
+information-theory limit (\CRLB).
+The only significant change is that the simple Gaussian likelihood must
+be replaced with the appropriate function.
+That change modifies the \CRLB\ calculation, and it also modifies the
+relevant data-analysis method, since cross-correlation (justified in
+\sectionname~\ref{s:methods}) is only appropriate when the noise is
+close to Gaussian.
+In a non-Gaussian world, the \RV\ estimation must proceed by explicit
+optimization of the likelihood (or Bayesian methods derived therefrom).
+
+The second way in which we might weaken our assumptions about the
+noise is that there might be contributions to the noise (especially
+those coming from unmodeled calibration artifacts) with unknown
+variance.
+That is, it is an optimistic assumption that we know the variance of
+the noise, and all the covariances.
+Our only comment about this problem is that, at least, the assumption
+is testable:
+It is possible with big observing programs to look at the empirical
+properties of the residuals away from the best-fit model.
+Do the residuals appear to have zero mean and reasonable empirical
+variance, given the assumed variance?
+These tests are like chi-squared tests; they test the completeness
+of the model.
+If the model is incomplete, it can be fixed by adding new components
+to the model (calibration vectors of various kinds, for example) or
+inflating estimates of noise variance. Or both.
 

### 2017-12-11 paper/paper.tex
decorate assumptions with commentary

@@
   calibrated near-perfectly \emph{consistently} across epochs (exposures).
+  We will break this assumption in our experiments, below.
 \item There are no residual telluric absorptions nor any residual sky
@@
   and removed.
-\item There is no time dependence (no epoch-to-epoch variations) in the
-  stellar spectrum.
+  We will also break this assumption in our experiments, below.
 \item The \emph{true} spectrum (that is, the latent spectrum that
@@
   known at better accuracy and precision than the main noise sources.
+  We will break this assumption too.
+  The whole point of this paper is to ask whether we can create a
+  good-enough proxy for the true spectrum.
+\item There is no time dependence (no epoch-to-epoch variations) in the
+  stellar spectrum.
 \item All noise sources contributing to the spectral data---which
@@
 
-...MOAR assumptions
-
-...limitations of our ``additional noise'' ideas and what we have considered.
-
 ...Hogg: insert some grand final statement.

### 2017-12-11 paper/paper.tex
tweaks

@@
 literature for maximum \RV\ precision (\todo{cite Butler, Bouchy,
-  Figueira, Lovis}) makes these same assumptions, plus additional \emph{even
-more restrictive assumptions}.
+  Figueira, Lovis}) makes all of these same assumptions,
+plus additional \emph{even more restrictive assumptions}.
 Under these disturbingly unrealistic assumptions, the \CRLB\ becomes

### 2017-12-11 paper/paper.tex
first shot at acks

@@
 
-...paragraph for each major result from the abstract
+...insert here one paragraph for each major result from the
+abstract. Each paragraph should emphasize the awesome of the result,
+but then discuss limitations or interpretation...
 
@@
 \acknowledgements
-MPIA hospitality, Montet, Foreman-Mackey, Bean, Stuermer, Rix, \HARPS\ team, ESO, Stars group meeting, Brewer
+It is a pleasure to thank
+  Jacob Bean (Chicago),
+  John Brewer (Yale),
+  Dan Foreman-Mackey (Flatiron),
+  Ben Montet (Chicago),
+  Hans-Walter Rix (MPIA),
+  Julian Stuermer (Chicago),
+and the participants in the weekly Stars Group Meeting at the Flatiron
+Institute for valuable contributions to this project.
+None of this would have been possible without the incredible effort
+of the \HARPS\ team and \ESO, who built the instrument and made the
+data used here public.
+MB thanks the Max-Planck-Institut f\"ur Astronomie in Heidelberg
+for hospitality during a critical phase of this project.
+
+\todo{add Hogg grant numbers}
 

### 2017-12-11 paper/paper.tex
notes to self

@@
 None of this would have been possible without the incredible effort
-of the \HARPS\ team and \ESO, who built the instrument and made the
+of the \HARPS\ team and \acronym{ESO}, who built the instrument and made the
 data used here public.
@@
 
-\todo{add Hogg grant numbers}
+\todo{...add Hogg grant numbers}
+
+\todo{...add facilities tags}
+
+\todo{...add software tags}
 

### 2017-12-17 paper/paper.tex
disc para

@@
 
-...insert here one paragraph for each major result from the
-abstract. Each paragraph should emphasize the awesome of the result,
-but then discuss limitations or interpretation...
+Fundamentally, what we have shown is that a data-driven model for
+stellar spectra can be accurate enough to deliver
+radial-velocity (\RV) measurements that saturate information-theoretic
+bounds.
+It is not surprising that an accurate model of a stellar spectrum
+yields good cross-correlation \RV\ measurements.
+What is interesting (at least to us) is that there is sufficient
+spectral information in a typical exoplanet-related \RV\ observing
+program to determine the mean stellar spectrum to sufficient
+quality that the \RV\ measurements are close to the best that can
+theoretically be made.
+It means that the information in the data that goes into determining
+the mean stellar spectrum does not come at the expense (in any sense)
+of the \RV\ information.
+
+...insert here one paragraph for each of the other major results from
+the abstract. Each paragraph should emphasize the awesome of the
+result, and/or discuss limitations or interpretation...
 

### 2017-12-17 paper/paper.tex
more of the same

@@
 
-Fundamentally, what we have shown is that a data-driven model for
+This \documentname\ reports good news for exoplanet-related
+radial-velocity (\RV) experiments.
+First and foremost, we have shown that a data-driven model for
 stellar spectra can be accurate enough to deliver
-radial-velocity (\RV) measurements that saturate information-theoretic
+\RV\ measurements that saturate information-theoretic
 bounds.
@@
 What is interesting (at least to us) is that there is sufficient
-spectral information in a typical exoplanet-related \RV\ observing
+spectral information in a typical observing
 program to determine the mean stellar spectrum to sufficient
@@
 of the \RV\ information.
+It also means that \RV\ experiments do not rely on having good models
+of stellar spectra; no physical model of a star or its photosphere
+is required.
 

### 2018-01-03 notebook/utils.py
cleaned up and added SNR scaling

@@ -1,0 +1,70 @@
+"""
+This code is part of the EPRV project.
+Copyright 2017,2018 David W Hogg (NYU) & Megan Bedell (Flatiron).
+"""
+
+import numpy as np
+
+c = 299792458. # m/s
+np.random.seed(42)
+sqrttwopi = np.sqrt(2. * np.pi)
+
+def doppler(rv):
+    beta = rv / c
+    return np.sqrt((1. - beta) / (1. + beta))
+
+def oned_gaussian(xs, mm, sig):
+    return np.exp(-0.5 * (xs - mm) ** 2 / sig ** 2) / (sqrttwopi * sig)
+
+def make_synth(rv, xs, ds, ms, sigs):
+    """
+    Generate a noiseless synthetic spectrum at given rv shift.
+    `rv`: radial velocity in m/s (or same units as `c` above)
+    `xs`: `[M]` array of wavelength values
+    `ds`: depths at line centers
+    `ms`: locations of the line centers in rest wavelength
+    `sigs`: Gaussian sigmas of lines
+    --
+    We take the view that lines multiply into the spectrum.
+    """
+    lnsynths = np.zeros_like(xs)
+    for d, m, sig in zip(ds, ms, sigs):
+        lnsynths += (d * oned_gaussian(xs * doppler(rv), m, sig))
+    return np.exp(lnsynths)
+
+def make_data(N, xs, ds, ms, sigs, snr):
+    """
+    Generate a set of N synthetic spectra.
+    `N`: number of spectra to make
+    `xs`: `[M]` array of wavelength values
+    `ds`: depth-like parameters for lines
+    `ms`: locations of the line centers in rest wavelength
+    `sigs`: Gaussian sigmas of lines
+    `snr`: desired SNR per pixel
+    """
+    M = len(xs)
+    data = np.zeros((N, M))
+    ivars = np.zeros((N, M))
+    rvs = 30000. * np.random.uniform(-1., 1., size=N) # 30 km/s bc Earth ; MAGIC
+    for n, rv in enumerate(rvs):
+        ivars[n, :] = snr**2.
+        data[n, :] = make_synth(rv, xs, ds, ms, sigs)
+        data[n, :] += np.random.normal(size=M) / np.sqrt(ivars[n, :])
+    return data, ivars, rvs
+    
+def dsynth_dv(rv, xs, ds, ms, sigs):
[…]

### 2018-01-03 paper/paper.tex
typo
single-line experiments
more work on demos

@@
 % text macros
+\newcommand{\documentname}{\textsl{Article}}
 \newcommand{\sectionname}{Section}
@@
 nothing here is controversial.
+We will come back to the scaling of uncertainty with various Gaussian line parameters in \sectionname~\ref{s:singleline}.
 
@@
 Discuss scaling of the CRLB with line FWHM, constrast/depth, etc.
-According to Lovis \& Fischer (2010) and Figueira (2017), $\sigma_{RV}$ should go as $\frac{\sqrt{FWHM}}{SNR \times C}$, where $C$ is the line contrast, or the depth of the line as a fraction of the continuum level \todo{(check on this definition of contrast)}.
+According to Lovis \& Fischer (2010) and Figueira (2017), $\sigma_{RV}$ should go as $\frac{\sqrt{FWHM}}{SNR \times C}$, where $C$ is the line contrast, or the depth of the line as a fraction of the continuum level.
 Do we see this?
@@
 Our next test involved a more realistic test spectrum.
-We measured the centers and depths of \todo{N} lines in the Solar spectrum between {x - y} $\ang$ and used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
+Using \HARPS\ spectra of the solar twin star HIP 54287, we measured the centers and depths of 12 lines in a 4-$\ang$ window around 5000 $\ang$.
+We used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
 
 What is the CRLB for this case?
-Binary mask cross-correlation is very far from saturating it.
-Cross-correlation or max-likelihood fitting with a perfect spectral template works much better.
+How close can we come to saturating it when using (a) a typical HARPS-like mask and (b) an optimized mask?
+In all cases, binary mask cross-correlation is very far from saturating the CRLB.
 Talk a little bit about why.
@@
 
-how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template. do a correct template and an incorrect template.
+how well does the mask/template approximate the real data?
+try fudging the relative line depths and locations believably, like maybe you have a slightly wrong spectral type or metallicity in the template.
+do a correct template and an incorrect template.
 

### 2018-01-03 notebook/utils.py

@@ -68,3 +68,108 @@ def calc_crlb(xs, ds, ms, sigs, N, ivars):
         crlbs[n] = np.sum(dmodel_dv * ivars[n, :] * dmodel_dv)
     crlb = 1. / np.sqrt(np.mean(crlbs))
     return crlb
+    
+def xcorr(data, model, ivars):
+    """
+    `data`: `[M]` array of pixels
+    `model`: `[M]` array of predicted pixels
+    `ivars`: `[M]` array of inverse variance values
+
+    Note: Presumes `ivars` is a list of diagonal entries; no
+      capacity for a dense inverse-variance matrix.
+    """
+    dd = data - np.mean(data)
+    mm = model - np.mean(model)
+    return np.sum(dd * ivars * mm) / np.sqrt(np.sum(dd * ivars * dd) * np.sum(mm * ivars * mm))
+
+def get_objective_on_grid(data, ivars, template, args, method, guess, halfwidth):
+    """
+    `data`: `[M]` array of pixel values
+    `ivars`: matched array of inverse variance values
+    `args`: list of inputs to go into `template` function after `rv`
+    `template`: function that makes the spectral template or mask
+    `method`: objective-function generator; currently `xcorr` or `chisq`
+    `guess`: where (in wavelength) to center the grid
+    `halfwidth`: half-width of the grid to make
+    """
+    drv = 25. # m/s grid size
+    rvs = np.arange(guess - halfwidth + 0.5 * drv, guess + halfwidth, drv) # m/s grid values
+    objs = np.zeros_like(rvs)
+    for i, rv in enumerate(rvs):
+        model = template(rv, *args)
+        objs[i] = method(data, model, ivars)
+    return rvs, objs
+
+def quadratic_max(xs, ys):
+    """
+    Find the maximum from a list, using a quadratic interpolation.
+    Note: REQUIRES (and `assert`s) that the xs grid is uniform.
+    """
+    delta = xs[1] - xs[0]
+    assert np.allclose(xs[1:] - xs[:-1], delta), "quadratic_max: not uniform grid!"
+    ii = np.nanargmax(ys)
+    if ii == 0:
+        print("quadratic_max: warning: grid edge")
+        return xs[ii]
+    if (ii + 1) == len(ys):
+        print("quadratic_max: warning: grid edge")
[…]

### 2018-07-18 paper/paper.tex
new title and abstract

--- 2017-12-17 paper/paper.tex
@@
 
-\documentclass[modern]{aastex61}
+\documentclass[modern]{aastex62}
 \usepackage{graphicx}
@@
 \setlength{\parindent}{1.4em} % trust in Hogg
-\shorttitle{achieving maximum radial-velocity precision}
+\shorttitle{data analysis for radial-velocity measurement}
 \shortauthors{bedell and hogg}
@@
 
-\title{Achieving maximum possible precision on stellar radial-velocity measurements in multi-epoch spectroscopy}
+\title{\textbf{Data analysis for maximum-precision radial-velocity mesurement}}
 
@@
 % context
-Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
-Despite the importance of these observations, the detailed computational methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
+Extremely precise stellar radial velocity (\RV) measurements are a
+foundational tool in the field of exoplanets, with hundreds of planets
+discovered by the Doppler method to date and many high-resolution
+spectrographs around the world dedicated to search and
+characterization.
+Current observations are delivering \RV\ measurements at $\ms$
+precision, while it is thought that some hardware systems ought to be
+delivering data capable of even higher precisions.
 % aims
-In this work, we look at sensible approaches to extracting \RV s from a data set of extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
+The question arises: Is there any precision loss in current
+data-analysis methods?
+That is, what data-analysis methods deliver \RV\ measurements that
+saturate the information-theoretic bound on precision---the
+Cram\'er--Rao bound (\CRLB)---given an extracted spectrum (and
+associated uncertainties) from a spectrograph?
 % methods
-We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
-We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
+We consider cross-correlation with a binary mask, a smoothed (or
+Gaussian-fit) cross-correlation with a binary mask, and
+cross-correlation with a stellar template.
+We consider both theoretically generated and data-driven templates,
+and look at some methods for data-driven telluric removal.
 % results
-We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
-We find that if there are micro-tellurics, they can be captured by a data-driven model.
-We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
+As expected, we find that cross-correlation with a correct stellar
+template produces \RV\ measurements that saturate the \CRLB.
+We find that pure cross-correlation with a binary mask under-performs,
+but the standard practice of post-processing the binary-mask result
+with a Gaussian fit does saturate the \CRLB, when the binary mask is
+``correct'' in a certain sense.
+As the template or mask becomes incorrect, \RV\ precision suffers.
+Most importantly, we find that using the data themselves to generate
+the stellar template does not cost any significant information; that
+is, a data-driven template will be correct enough to saturate the
+\CRLB, provided that there are multiple epochs of observations and the
+star does not vary substantially between epochs.
+As a side note, we find that tiny telluric features can affect \RV\ precision
+substantially, but that they can be modeled and removed by a data-driven
+method, provided that there is sufficient barycentric velocity variation.
 \end{abstract}
@@
 \section{Introduction}
+
+\todo{needs reduction in scope, perhaps?}
 

### 2018-07-18 paper/paper.tex
tweaks

@@
 That is, what data-analysis methods deliver \RV\ measurements that
-saturate the information-theoretic bound on precision---the
-Cram\'er--Rao bound (\CRLB)---given an extracted spectrum (and
+saturate the information-theoretic 
+Cram\'er--Rao bound (\CRLB) on precision, given an extracted spectrum (and
 associated uncertainties) from a spectrograph?
@@
 substantially, but that they can be modeled and removed by a data-driven
-method, provided that there is sufficient barycentric velocity variation.
+method, provided that there is sufficient barycentric velocity variation to
+separate the telluric and star rest frames.
 \end{abstract}

### 2018-10-29 variability/var.tex
start

--- 2018-07-18 paper/paper.tex
@@
+% var.tex : a paper about EPRV in the presence of spectral variability
 % This file is part of the EPRV project.
-% Copyright 2017, 2018 the authors.
+% Copyright 2018 the authors.
 
 % Style notes
+% -----------
 % - Re-define acronyms in each section? So less flippy flippy?
+
+% To-do items
+% -----------
+% - figure out what experiments to do and do them?
+% - Draft it.
+% - Get a 6th keyword; re-consider keywords.
 
@@
 \setlength{\parindent}{1.4em} % trust in Hogg
-\shorttitle{data analysis for radial-velocity measurement}
-\shortauthors{bedell and hogg}
+\shorttitle{precise radial velocities for variable stars}
+\shortauthors{hogg and others}
 
@@
 
-\title{\textbf{Data analysis for maximum-precision radial-velocity mesurement}}
-
-\author[0000-0001-9907-7742]{Megan Bedell}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affil{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
+\title{\textbf{Precise radial-velocity measurements in the presence of stellar variability}}
 
@@
 % context
-Extremely precise stellar radial velocity (\RV) measurements are a
-foundational tool in the field of exoplanets, with hundreds of planets
-discovered by the Doppler method to date and many high-resolution
-spectrographs around the world dedicated to search and
-characterization.
-Current observations are delivering \RV\ measurements at $\ms$
-precision, while it is thought that some hardware systems ought to be
-delivering data capable of even higher precisions.
+foo.
 % aims
-The question arises: Is there any precision loss in current
-data-analysis methods?
-That is, what data-analysis methods deliver \RV\ measurements that
-saturate the information-theoretic bound on precision---the
-Cram\'er--Rao bound (\CRLB)---given an extracted spectrum (and
-associated uncertainties) from a spectrograph?
+bar.
 % methods
-We consider cross-correlation with a binary mask, a smoothed (or
-Gaussian-fit) cross-correlation with a binary mask, and
-cross-correlation with a stellar template.
-We consider both theoretically generated and data-driven templates,
-and look at some methods for data-driven telluric removal.
+blah.
 % results
-As expected, we find that cross-correlation with a correct stellar
-template produces \RV\ measurements that saturate the \CRLB.
-We find that pure cross-correlation with a binary mask under-performs,
-but the standard practice of post-processing the binary-mask result
-with a Gaussian fit does saturate the \CRLB, when the binary mask is
-``correct'' in a certain sense.
-As the template or mask becomes incorrect, \RV\ precision suffers.
-Most importantly, we find that using the data themselves to generate
-the stellar template does not cost any significant information; that
-is, a data-driven template will be correct enough to saturate the
-\CRLB, provided that there are multiple epochs of observations and the
-star does not vary substantially between epochs.
-As a side note, we find that tiny telluric features can affect \RV\ precision
-substantially, but that they can be modeled and removed by a data-driven
-method, provided that there is sufficient barycentric velocity variation.
+duh.
 \end{abstract}
@@
 \keywords{
-atmospheric effects
----
+% atmospheric effects
+% ---
 binaries: spectroscopic
@@
 
-\todo{needs reduction in scope, perhaps?}
-
-Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from Galactic dynamics to large-scale structure and cosmic expansion.
-Fundamentally, accuracy of radial-velocity mesurement is limited by the quality of spectral models of stars, or templates.
-In recent decades, however, a new focus has emerged on \emph{extremely precise} but purely \emph{relative} \RV\ measurements of stars as a means of detecting exoplanets and other kinds of companions.
-For this application, measuring the absolute velocity of the star is much less important than resolving changes through time.
-With sub-$\ms$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
-At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
-In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision \RV s: the method of analysis used to estimate the \RV\ given an extracted stellar spectrum.
-
-While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of reference spectrum or template to the treatment of telluric absorption features.
-Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
-Queloz1995, Baranne1996).
-A lesser-used alternative is cross-correlation or maximum-likelihood fitting using a template spectrum generated from the data (Buchhave, Anglada-Escude). 
-This often involves masking or downweighting of the most telluric-contaminated spectral regions.
-
-On the other hand, gas-cell-calibrated spectrographs like \HIRES\ commonly adopt a more comprehensive forward-modeling approach. 
-In this framework, the instrumental line spread function, the stellar RV, and other parameters are optimized by modeling each observation as a combination of shifted and instrument-convolved template spectra for the star and gas cell.
-[say a bit more]
-
-Our primary aim in this work is to establish a common frame of reference for these methods based on their relative ability to extract maximum-precision RVs and their robustness in the face of various expected structured noise sources.
-
-For the purposes of this work, we assume that our data set consists of multiple observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
-This means we are not considering the case of a real-time data reduction pipeline. 
-We also assume a \HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. 
-We return to this point in the discussion \todo{(do this)}.
-
-We begin in Section \ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum. 
-In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting. 
-We then implement these techniques to extract \RV s using 5-\ang-wide sections of simulated and real spectra, as described in Section \ref{s:data}. 
-We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
-
-\section{Information Theory}
-\label{s:info}
-
-Given finite, noisy data, there is a limit coming from information
-theory on how well any parameter of interest can be measured.
-This limit is
-sometimes known as the Cram\'er--Rao Lower Bound (\CRLB; \todo{cite CR and wikipedia}).
-In its simplest form, it constrains the variance of any unbiased
-frequentist estimator.
-Estimators can only beat this bound by taking on bias.
-
-This bound is strong, but it cannot be calculated \emph{objectively};
-any description of the information in a data set requires
-subjective decisions (that is, debatable decisions), which are inputs
-to a model for the generation of the data.
-In particular, information-theory bounds require a model for the
-probability of the data given the parameter of interest; that is, they
-require a likelihood function.
-As an aside: Bayesians need a likelihood function for inference, and
-frequentists need a likelihood function for analyzing their estimators.
-The likelihood function is therefore central to all statistical methods,
-and the core component of any data analysis.
-
-In particular, the \CRLB\ is related to the Fisher Information that is
-frequently used in cosmological experiment forecasts (\todo{CITE
-  SHIT}).
-It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
-estimator (that is, the information in the velocity estimator) must be smaller than the
-second derivative of the \emph{negative} log likelihood function:
-\begin{equation}
-\frac{1}{\sigma_v^2} \leq E\left[\frac{\dd^2}{\dd v^2}(-\ln L)\right] \quad,
-\end{equation}
-where $L$ is the likelihood, or the probability of the data given the velocity,
-and the expectation $E[\cdot]$ is taken over all possible data (that is, it is an
-expectation under the likelihood).
-If there are many other nuisance parameters, a Bayesian can marginalize them
-out to make $L$ the marginalized likelihood for the velocity, or a frequentist
-can include them in the estimation and compute an inverse covariance matrix $C^{-1}$
-for the full parameter vector, which will have the bound
-\begin{equation}
-v^T\cdot C^{-1}\cdot v \leq v^T\cdot Q\cdot v
-\end{equation}
-\begin{equation}
-Q_{ij} \equiv E\left[\frac{\dd^2}{\dd\theta_i\,\dd\theta_j}(-\ln L)\right] \quad,
-\end{equation}
-where $Q$ is the information tensor, $i,j$ are indices into that
-tensor, corresponding parameters are $\theta_i, \theta_j$, and $v$ is
-any arbitrary vector.  That is, the bound is not just on every
-parameter, but really on on any combination of parameters.
-
-In the case of extreme-precision \RV, this says that any computation of the limits
-on estimating stellar velocities (or stellar velocity changes)
-requires a likelihood function.
-This, in turn, requires making assumptions about the stellar spectrum (the
-latent, unobserved, \emph{true} spectrum that could only be observed
-with arbitrarily good data) and the noise processes that distort the
-spectrum in any finite, noisy observation.
-These noise processes ought (in principle) to include any beliefs about
-unsubtracted telluric absorptions or emissions, and time variability
-of the star itself, not to mention photon and spectrograph read and
-extraction noise.
-However---and in keeping with the literature---we can make maximal or
-best-case assumptions, and ask what we would get for \RV\ precision
-in that best case.
-Then we can ask (with fake data) how much different choices about our
-data analysis, and different adversarial injections of noise,
-tellurics, spectrograph issues, and other wrongness can prevent us
-from achieving the best-case bounds.
-
-The most optimistic assumptions we can possibly make are the following.
-\begin{itemize}
-\item The spectrograph is perfectly calibrated in a wavelength sense.
-\item The spectrograph is consistently and repeatably calibrated in a
-  flux sense. That is, the spectrum does not need to be perfectly
-  continuum-normalized or perfectly flux-normalized, but it is
-  calibrated near-perfectly \emph{consistently} across epochs (exposures).
-  We will break this assumption in our experiments, below.
-\item There are no residual telluric absorptions nor any residual sky
-  emission in the spectra. That is, the sky is near-perfectly calibrated
-  and removed.
-  We will also break this assumption in our experiments, below.
-\item The \emph{true} spectrum (that is, the latent spectrum that
-  would be observed if the data were far, far better) of the star is
-  known at better accuracy and precision than the main noise sources.
-  We will break this assumption too.
-  The whole point of this paper is to ask whether we can create a
-  good-enough proxy for the true spectrum.
-\item There is no time dependence (no epoch-to-epoch variations) in the
-  stellar spectrum.
-\item All noise sources contributing to the spectral data---which
-  include at a minimum photon noise, read noise, residual telluric
-  issues, residual wavelength-calibration issues, and residual stellar
-  spectral variability---are, when summed together, indistinguishable
-  from being zero-mean, Gaussian, and additive.
-\item The complete variance tensor of the total noise on the spectrum is known, and
-  known correctly. That is, the variance of the noise at every pixel
-  is known, as is the covariance between nearby pixels, and in the
-  same \emph{true} sense that the true spectrum of the star is known.
-\end{itemize}
-It is worthy of note that every formula (that we know) in the
-literature for maximum \RV\ precision (\todo{cite Butler, Bouchy,
-  Figueira, Lovis}) makes all of these same assumptions,
-plus additional \emph{even more restrictive assumptions}.
-Under these disturbingly unrealistic assumptions, the \CRLB\ becomes
-\begin{equation}\label{eq:crlb}
-\frac{1}{\sigma_v^2} \leq \left[\frac{\dd f}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f}{\dd v}\right]
-\end{equation}
-where the derivatives are of the true spectrum with respect to
-velocity (which can be thought of as the derivative of a Doppler
-operator with respect to velocity acting on the true spectrum), the
-derivatives are column vectors, and $C^{-1}$ is the inverse of the covariance matrix
-describing the true variance and covariances of the noise process.
-
-Formulae (bounds) related to (\ref{eq:crlb}) have appeared previously
-in the literature, although in somewhat more complicated forms.
-Butler (\todo{cite}) shows a similar expression, but makes the additional
-assumption that every pixel is independent; that is, that there is no
-covariance between nearby pixels from, say, extraction or continuum normalization.
-Bouchy (\todo{cite}) shows an expression that assumes that the pixels
-are independent and that the noise is dominated by a combination of photon
-and read noise. 
-That work also replaces the model spectrum with a set of optimal
-weights for cross-correlation, which amounts to very much the same thing.
-Figueira (\todo{cite}) and \todo{others (cite)} show a per-line
-bound, based on assumptions similar to those of Bouchy but with the additional assumption that
-each line is well approximated by a Gaussian (squared exponential) in shape.
-These literature bounds are all extremely similar, and similar to (\ref{eq:crlb});
-nothing here is controversial.
-
-Now that we have established the bound, how can we saturate it? That is,
-how can we make measurements (estimators)  with variances that come close to the \CRLB?
-The short answer is that reliable bound-saturating estimators are maximum-likelihood estimators.
-The long answer is long, but part of it is that the bound can only be saturated if the
-assumptions are correct, and note that the assumptions include that we know the
-\emph{true} spectrum of the star, among many other things.
-That is, if we can take data that come close to meeting the assumptions, and
-if we can find a spectral model that comes close to the true spectrum of the star,
-then in principle we can estimate radial velocities with a precision close to the \CRLB.
-
-Given that computational stellar spectral models have physical issues, the best way
-to get an accurate estimate of the true spectrum of the star is going to be data-driven.
-In what follows, we are going to show that---for a typical \HARPS-like observing program---%
-it is possible to build a data-driven model of the stellar spectrum that is good enough
-to nearly-saturate the \CRLB.
-That is, with a data-driven model of the true stellar spectrum, model inaccuracy will
-not dominate the \RV\ noise budget.
-We will additionally bring evidence that telluric features can be a significant problem,
-but that they also can plausibly be mitigated with a data-driven model.
-
-\section{Radial-velocity estimators}
-\label{s:methods}
-
-Most methods for radial-velocity estimation involve some kind of
-cross-correlation of the data with a template.
-This might be surprising, since (as we noted above) bound-saturating
-estimators will be maximum-likelihood estimators.
-However, under some simple assumptions, cross-correlation optima are
-also maximum-likelihood optima.
-
-For example, when the likelihood is Gaussian, the log-likelihood
-contains a chi-squared-like term
-\begin{equation}
--\frac{1}{2}\,[y - f]\T\cdot C^{-1}\cdot [y - f] \quad ,
-\end{equation}
-where $y$ is a column vector of the data, $f$ is the model spectrum,
-and $C$ is the covariance matrix describing the noise variance.
-In the case of independent pixel noise this just becomes an
-inverse-variance weighted sum of squared residuals (chi-squared).
-If we expand this quadratic form, it becomes
-\begin{equation}
-{}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f 
--\frac{1}{2}\,f\T\cdot C^{-1}\cdot f \quad.
-\end{equation}
-That is, the chi-squared term factors into a term that is the square
-of the model, a term that is the square of the data, and a term that
-is a cross-correlation.
-As long as the amplitude of $f$ doesn't depend on velocity, and the
-$C$ tensor is ``smooth'' in certain respects, the dependence of this
-chi-squared term on velocity will be dominated by the
-cross-correlation.
-For this reason, we will use a variant of cross-correlation
-optimization as our radial-velocity estimator.
-
-...In detail, the cross-correlation we will optimize is...
-
-A variety of choices exist for the cross-correlation template $f$. A few possibilities include:
-\begin{itemize}
-\item a ``binary'' mask, or a spectrum of zero value with non-zero peaks at the expected rest-frame wavelengths of the spectral lines of interest.
-A cross-correlation between this mask and the observed spectrum will effectively compute the average line profile with its minimum corresponding to the best-fit RV.
-This method has the benefit of trivially excluding telluric-contaminated or otherwise unreliable regions of the spectrum by deleting them from the mask.
-A desired relative weighting of lines can be incorporated by setting the non-zero mask values to different amplitudes at different lines (cite Pepe).
-This method is currently used by the \HARPS\ pipeline (cite).
-\item synthetic spectrum (i.e.
-a slightly wrong model prediction, since the stellar model will never be a perfect reproduction of the star in question): sub-tweet: some surveys cross-correlate with a single (same) template for every single star, which is extremely wrong.
-you can be more right by using a spectral-type-specific template.
-\item data-driven template spectrum: some surveys (HIRES) use a single observation as a template. combining information from many visits is better because the combined SNR is amazing.
-\end{itemize}
-
-One final implementation detail is the following:
-When we optimize the cross-correlation function, we don't run an optimizer;
-we perform brute-force search on a linear velocity grid.
-This grid, however, does not need to be fine.
-We use a coarse grid (\todo{give grid spacing here}), which does not
-over-resolve the velocity peak at the optimum.
-We then take the highest value on the grid and the values on either
-side, fit a parabola through those three values, and take as the optimal
-velocity the peak of that parabola.
-That is, we obtain sub-grid velocity resolution by means of a quadratic
-interpolation at the peak on the finite grid.
-\todo{Do we need to demonstrate this with a toy figure?}
-
-\begin{figure}
-\centering
-% \includegraphics[width=\columnwidth]{binarymask}
-\caption{Illustration of a weighted binary mask (blue windows) plotted over a small segment of the solar spectrum (black).
-The vertical length of each mask window corresponds to the weight given to the line when performing a cross-correlation between mask and spectrum.}
-\label{fig:binarymask}
-\end{figure}
-
-\section{Data}
-\label{s:data}
-
-The experiments we perform in this paper primarily use artificial data, for the simple reason that we require access to the underlying true RVs in order to assess the performance of our methods. 
-We attempt to generate the fake data as realistically as possible for the case of a high-resolution, high-signal-to-noise spectrum of a Sun-like star taken with \HARPS. 
-With these data, we are able to compute the \CRLB.  
-We can also isolate the effects of various confounding noise sources like telluric features and imperfect continuum normalization by injecting them into the artificial data, which is generated free of correlated noise by default.
-
-We will also show tests on real \HARPS\ data for two stars, one hosting a companion and
-one (apparently) companion-free (at least at our precision). 
-These tests serve as a demonstration that the methods presented here are useful when applied to real data and the assumptions made in our artificial data generation are realistic.
-
-\subsection{Artificial Data}
-
-The basic model that we use to generate fake spectra is extremely simple, assuming a perfectly normalized continuum; isolated, perfectly Gaussian spectral lines; and white noise only. 
-We simulated a 5-\ang\ region with random noise corresponding to an SNR of 100, a typical value for a single \RV\ spectrum. 
-Lines were inserted as Gaussians with prescribed centers, widths, and depths. 
-Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 \ang. 
-In the simplest test case, an arbitrary single line at 5000 \ang\ was used (see Section \ref{s:singleline}). 
-However, for most of our tests we injected a set of 12 lines with centers between 4998 - 5002 \ang. 
-The Gaussian parameters for these lines were based on fits to a reflected Solar spectrum from Vesta taken by \HARPS. \todo{(find citation/program number)} 
-The resulting spectrum strongly resembles a 5-\ang\ chunk taken from a typical \HARPS\ spectrum of a Sun-like star (Figure \ref{fig:spectra}).
-
-We repeated this synthesis to make a set of \todo{64} spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion).
-The \RV s were applied as a multiplicative Doppler shift on the central wavelength of each spectral line to be injected. 
-At this point, we are able to run a cross-correlation analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.
-
-\begin{figure}
-\centering
-% \includegraphics[width=\columnwidth]{spectra}
-\caption{}
-\label{fig:spectra}
-\end{figure}
-
-\subsection{\HARPS\ Data}
-
-munging notes
-
-reference to figure comparing real \& fake data
-
-\section{Experiments and results}
-\label{s:experiments}
-
-\subsection{Single-Line}
-\label{s:singleline}
-
-We began with the simplest test case possible: a single, strong absorption feature.
-
-What is the CRLB?
-
-Discuss scaling of the CRLB with line FWHM, constrast/depth, etc.
-According to Lovis \& Fischer (2010) and Figueira (2017), $\sigma_{RV}$ should go as $\frac{\sqrt{FWHM}}{SNR \times C}$, where $C$ is the line contrast, or the depth of the line as a fraction of the continuum level \todo{(check on this definition of contrast)}.
-Do we see this?
-
-\subsection{Binary Mask}
-
-Our next test involved a more realistic test spectrum.
-We measured the centers and depths of \todo{N} lines in the Solar spectrum between {x - y} $\ang$ and used these measurements to generate a synthetic spectrum that closely resembles a typical observation of a Sun-like star.
-
-What is the CRLB for this case?
-Binary mask cross-correlation is very far from saturating it.
-Cross-correlation or max-likelihood fitting with a perfect spectral template works much better.
-Talk a little bit about why.
-
-\subsection{Rigid Template}
-
-how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template. do a correct template and an incorrect template.
-
-even a tiny wrongness to the template is visible in the RV variance.
-
-\subsection{Data-Driven Template}
-
-how do we construct the template? show that it comes close to saturating the CRLB.
-
-comment: there are far better ways to derive the data-driven template. we leave this to future work.
-
-\subsection{Realistic Sources of Structured Noise}
-
-additional noise considerations:
-\begin{itemize}
-\item (micro-) telluric features: tellurics that you can't identify are important
-\item subtle continuum trends: not that big a deal
-\item cosmic rays?: we might skip this and add to discussion
-\item additive signals?: for example sky emission, lunar and other reflected solar; we might skip this and just add to discussion
-\end{itemize}
-
-\subsection{Real Data}
-
-all the data-driven code works on real data.
-we get a better RMS on the star from the data-driven template than from the binary mask.
-based on this, we believe that the \HARPS\ pipeline doesn't strictly use a binary mask.
-
-\section{Summary and discussion}
-\label{s:summary}
-
-This \documentname\ reports good news for exoplanet-related
-radial-velocity (\RV) experiments.
-First and foremost, we have shown that a data-driven model for
-stellar spectra can be accurate enough to deliver
-\RV\ measurements that saturate information-theoretic
-bounds.
-It is not surprising that an accurate model of a stellar spectrum
-yields good cross-correlation \RV\ measurements.
-What is interesting (at least to us) is that there is sufficient
-spectral information in a typical observing
-program to determine the mean stellar spectrum to sufficient
-quality that the \RV\ measurements are close to the best that can
-theoretically be made.
-It means that the information in the data that goes into determining
-the mean stellar spectrum does not come at the expense (in any sense)
-of the \RV\ information.
-It also means that \RV\ experiments do not rely on having good models
-of stellar spectra; no physical model of a star or its photosphere
-is required.
-
-...insert here one paragraph for each of the other major results from
-the abstract. Each paragraph should emphasize the awesome of the
-result, and/or discuss limitations or interpretation...
-
-In \sectionname~\ref{s:info} we listed some exceedingly
-optimistic assumptions under which we could compute the
-information-theoretic limit on radial-velocity precision.
-Some of these assumptions are worthy of discussion: How true are they,
-and what happens, both from an information-theory perspective and from
-a methodological perspective, when we relax them?
-
-One of these assumptions, for example, is that the true stellar
-spectrum is constant with time; that is, the variations from
-epoch-to-epoch are from noise or nuisances (like telluric absorption).
-This assumption is critical and deep.
-The entire edifice of extreme-precision \RV\ is built on this assumption:
-With imperfect models of stellar structure and photospheres, it is
-impossible to know the \emph{absolute} \RV\ of a star to high precision;
-we satisfy ourselves that we can measure \emph{relative} velocities
-to arbitrarily high precision.
-Relative precision---and everything in this paper---depends on our
-ability to measure shifts from epoch to epoch.
-If the stellar spectrum is permitted (or observed) to change from epoch
-to epoch, there is no fundamental way to determine or confirm spectral
-shifts corresponding to Doppler shifts.
-That is, even relative \RV\ measurements depend on a constancy of the
-stellar spectrum.
-This is a deep issue, extending far beyond the scope of this paper.
-Time-variable stellar spectra are not killing to the extreme-precision
-\RV\ program!
-But at current practices they may be limiting precision,
-and to ameliorate their effects will require substantial and non-trivial adjustments to the
-\foreign{status quo}.
-We hope to address these in future work.
-
-We also assumed that the noise is zero-mean, Gaussian, and with known
-variance tensor.
-There are certainly sources of noise that might violate these
-assumptions.
-There are two ways in which we might weaken our assumptions about
-the noise.
-The first is that the total noise acting on the system---and recall
-that noise for us is anything not accounted for in our model of the
-data---might not be Gaussian in form.
-This will happen if the photon counts (per pixel) get low.
-But it can also happen if there are unmodeled calibration artifacts
-that don't happen to have Gaussian-distributed amplitudes.
-This is not a problem in principle with computing or saturating the
-information-theory limit (\CRLB).
-The only significant change is that the simple Gaussian likelihood must
-be replaced with the appropriate function.
-That change modifies the \CRLB\ calculation, and it also modifies the
-relevant data-analysis method, since cross-correlation (justified in
-\sectionname~\ref{s:methods}) is only appropriate when the noise is
-close to Gaussian.
-In a non-Gaussian world, the \RV\ estimation must proceed by explicit
-optimization of the likelihood (or Bayesian methods derived therefrom).
-
-The second way in which we might weaken our assumptions about the
-noise is that there might be contributions to the noise (especially
-those coming from unmodeled calibration artifacts) with unknown
-variance.
-That is, it is an optimistic assumption that we know the variance of
-the noise, and all the covariances.
-Our only comment about this problem is that, at least, the assumption
-is testable:
-It is possible with big observing programs to look at the empirical
-properties of the residuals away from the best-fit model.
-Do the residuals appear to have zero mean and reasonable empirical
-variance, given the assumed variance?
-These tests are like chi-squared tests; they test the completeness
-of the model.
-If the model is incomplete, it can be fixed by adding new components
-to the model (calibration vectors of various kinds, for example) or
-inflating estimates of noise variance. Or both.
-
-...Hogg: insert some grand final statement.
+Hello World!
 
@@
 It is a pleasure to thank
-  Jacob Bean (Chicago),
-  John Brewer (Yale),
-  Dan Foreman-Mackey (Flatiron),
-  Ben Montet (Chicago),
-  Hans-Walter Rix (MPIA),
-  Julian Stuermer (Chicago),
+  Ben Pope (Flatiron)
 and the participants in the weekly Stars Group Meeting at the Flatiron
 Institute for valuable contributions to this project.
-None of this would have been possible without the incredible effort
-of the \HARPS\ team and \acronym{ESO}, who built the instrument and made the
-data used here public.
-MB thanks the Max-Planck-Institut f\"ur Astronomie in Heidelberg
-for hospitality during a critical phase of this project.
 

### 2018-10-29 variability/var.tex
title and abstract

@@
 \author[0000-0003-2866-9403]{David W. Hogg}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
+\affil{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
@@
 
+\author{Others}
+
 \begin{abstract}\noindent
 % context
-foo.
+Spectrographs that deliver the very most precise radial-velocity (RV)
+measurements of stars---for the purpose of detecting exoplanets---are
+obtaining empirical root-variances of around $1\,\ms$.
+This is impressive, but not as precise as information theory
+permits.
+One possible source of variance is stellar variability, from stellar
+surface convection, sunspots, and activity.
+Spectral variability is undeniably present at some level for all
+convective-surface stars, and it undermines the fundamental
+assumptions of essentially all extreme-precision methodologies at
+present.
 % aims
-bar.
+Here we develop extreme-precision (EPRV) estimators that
+deliver RV precision without assuming that the stellar spectrum is
+constant in time.
+These methods are designed to work in pathological cases, where the
+variability is large, so that they will perform extremely well in more
+realistic situations in which the variability is small.
 % methods
-blah.
+They are probabilistic models, built on two important assumptions:
+The first is that the stellar variability is somehow compact, either
+in dimensionality or in total variance, or both.
+The second is that---in the limit of large data---the stellar
+variability is uncorrelated with the RV signals from any planets
+orbiting the star.
 % results
-duh.
+We find that, given sufficient data, our methods converge to saturating
+information-theoretic RV bounds, provided that the assumptions hold.
+This methodology opens up the possibility of reducing measurement variances,
+and widening the set of stars available for EPRV study.
 \end{abstract}
@@
 
+\section*{~}
+\clearpage
 \section{Introduction}
@@
 It is a pleasure to thank
-  Ben Pope (Flatiron)
+  Megan Bedell (Flatiron),
+  Benjamin Pope (NYU),
 and the participants in the weekly Stars Group Meeting at the Flatiron

### 2019-06-06 paper/paper.tex
letter
title
abstract, baby

--- 2018-07-18 paper/paper.tex
@@
 % This file is part of the EPRV project.
-% Copyright 2017, 2018 the authors.
+% Copyright 2017, 2018, 2019 the authors.
 
@@
 
-\title{\textbf{Data analysis for maximum-precision radial-velocity mesurement}}
+\title{\textbf{How to measure maximally precise radial velocities from a set of spectra:\\
+               The time-independent spectrum case}}
 
@@
 Current observations are delivering \RV\ measurements at $\ms$
-precision, while it is thought that some hardware systems ought to be
-delivering data capable of even higher precisions.
+precision
+It appears that many hardware systems ought to be---on the basis of
+information-theoretic considerations---capable of delivering even
+higher precisions.
 % aims
-The question arises: Is there any precision loss in current
-data-analysis methods?
-That is, what data-analysis methods deliver \RV\ measurements that
-saturate the information-theoretic bound on precision---the
-Cram\'er--Rao bound (\CRLB)---given an extracted spectrum (and
+Here we ask: What data-analysis methods deliver \RV\ measurements that
+saturate the
+Cram\'er--Rao bound (\CRLB) on precision, given an extracted spectrum (and
 associated uncertainties) from a spectrograph?
+We presume that the stellar spectrum does not vary with time (this is
+incorrect in detail).
 % methods
-We consider cross-correlation with a binary mask, a smoothed (or
-Gaussian-fit) cross-correlation with a binary mask, and
-cross-correlation with a stellar template.
-We consider both theoretically generated and data-driven templates,
-and look at some methods for data-driven telluric removal.
+We consider a smoothed (or
+Gaussian-fit) cross-correlation with a binary mask,
+cross-correlation with a stellar template,
+and maximum-likelihood methods.
+We draw out the relationships and equivalencies across these methods.
+We consider both theoretically generated and data-driven templates.
 % results
-As expected, we find that cross-correlation with a correct stellar
+We find (as expected) that cross-correlation with a correct stellar
 template produces \RV\ measurements that saturate the \CRLB.
-We find that pure cross-correlation with a binary mask under-performs,
-but the standard practice of post-processing the binary-mask result
-with a Gaussian fit does saturate the \CRLB, when the binary mask is
+We also find that 
+the standard practice of fitting a Gaussian to the cross-correlation
+with a binary mask saturates the \CRLB, so long as the binary mask is
 ``correct'' in a certain sense.
 As the template or mask becomes incorrect, \RV\ precision suffers.
+Un-corrected telluric absorption lines can also harm \RV\ precision.
 Most importantly, we find that using the data themselves to generate
-the stellar template does not cost any significant information; that
-is, a data-driven template will be correct enough to saturate the
-\CRLB, provided that there are multiple epochs of observations and the
-star does not vary substantially between epochs.
-As a side note, we find that tiny telluric features can affect \RV\ precision
-substantially, but that they can be modeled and removed by a data-driven
-method, provided that there is sufficient barycentric velocity variation.
+the stellar and telluric templates does not cost any significant
+information.
+That is, data-driven stellar and telluric templates can be correct
+enough to saturate the \CRLB, provided that there are multiple epochs
+of observations and the star does not vary substantially between
+epochs.
+If the star does vary substantially, a lot more needs to be known
+before the relevant \CRLB\ even can be computed.
 \end{abstract}

### 2019-06-11 paper/paper.tex
abstract

@@
 \newcommand{\m}{\unit{m}}
+\newcommand{\cm}{\unit{cm}}
 \newcommand{\s}{\unit{s}}
 \newcommand{\kms}{\km\,\s^{-1}}
-\newcommand{\ms}{\m\,\s^{-1}}
+\newcommand{\mps}{\m\,\s^{-1}}
+\newcommand{\cmps}{\cm\,\s^{-1}}
 \newcommand{\ang}{\text{\normalfont\AA}}
@@
 characterization.
-Current observations are delivering \RV\ measurements at $\ms$
-precision
+Current observations are delivering \RV\ measurements at $\mps$
+precision.
 It appears that many hardware systems ought to be---on the basis of
-information-theoretic considerations---capable of delivering even
-higher precisions.
+both spectrograph-stability and information-theoretic
+considerations---capable of delivering even higher precisions.
 % aims
 Here we ask: What data-analysis methods deliver \RV\ measurements that
-saturate the
-Cram\'er--Rao bound (\CRLB) on precision, given an extracted spectrum (and
-associated uncertainties) from a spectrograph?
+saturate the Cram\'er--Rao bound (\CRLB) on precision, given an
+extracted spectrum (and associated uncertainties) from a spectrograph?
 We presume that the stellar spectrum does not vary with time (this is
@@
 For this application, measuring the absolute velocity of the star is much less important than resolving changes through time.
-With sub-$\ms$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
+With sub-$\mps$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
 At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.

### 2019-06-11 paper/paper.tex
intro

@@
 % Style notes
-% - Re-define acronyms in each section? So less flippy flippy?
+% -----------
+% - Re-define acronyms in each section? So less flippy flippy? Yes that's the thing.?
+% -
+
+% To-do items
+% -----------
+% - search for all \todo
+% - search for all HOGG
+% - search for all BEDELL
+% - search for all CITE
+% - should we fun-cite some ancient papers in the first paragraph, like Hubble and Oort?
 
@@
 \newcommand{\project}[1]{\textsl{#1}}
-\newcommand{\foreign}[1]{\textsl{#1}}
 \newcommand{\HARPS}{\project{\acronym{HARPS}}}
@@
 \newcommand{\CRLB}{\acronym{CRLB}}
+\newcommand{\foreign}[1]{\textsl{#1}}
 
@@
 
-\todo{needs reduction in scope, perhaps?}
+It is ridiculous to open this \documentname\ by pointing out that
+the spectroscopic measurement of radial velocity (\RV) has been a
+critical tool across many subfields of astronomy, from Galactic
+dynamics to large-scale structure and cosmic expansion.
+Fundamentally, the accuracy of radial-velocity mesurement is limited
+by the quality of the spectral models of the targets of the
+spectroscopy.
+In recent decades, however, a new focus has emerged on \emph{extremely
+  precise} but \emph{purely relative} \RV\ measurements of stars as a
+means of detecting exoplanets and other kinds of companions.
+That is, we are looking for \emph{precision, not accuracy}.
+For this application, measuring the absolute velocity of the star is
+much less important than resolving changes through time.
+This is the problem we address in this \documentname.
 
-Accurate spectroscopic measurements of radial velocities (\RV s) have long been a critical tool across many subfields of astronomy, from Galactic dynamics to large-scale structure and cosmic expansion.
-Fundamentally, accuracy of radial-velocity mesurement is limited by the quality of spectral models of stars, or templates.
-In recent decades, however, a new focus has emerged on \emph{extremely precise} but purely \emph{relative} \RV\ measurements of stars as a means of detecting exoplanets and other kinds of companions.
-For this application, measuring the absolute velocity of the star is much less important than resolving changes through time.
-With sub-$\mps$-level precision required for the detection of Earth-like planets, vast amounts of effort have been poured into the engineering challenges of constructing a sufficiently stable spectrograph (or, failing that, being able to track the drift of a spectrograph over time) \todo{(cite some review papers)}.
-At the same time, great strides have been made towards disentangling true Doppler shifts from imitative signals like the spectral signatures of stellar pulsations and magnetic activity features \todo{(cite some review papers)}.
-In this work, we investigate another fundamentally important but less commonly discussed aspect of the quest for maximum-precision \RV s: the method of analysis used to estimate the \RV\ given an extracted stellar spectrum.
+Earth-like planets induce $0.1\,\mps$-level signals in their host
+stars!
+Substantial hardware, software, and systems-engineering effort has
+been put into the challenges of making measurements at this level of
+precision (CITE some reviews).
+It appears that contemporary spectrographs are stable at the
+$0.05\,\mps$ level (CITE THINGS), and that the information content in
+the data---which is set by the spectral features, the spectrograph
+resolution, the wavelength range, and the observational
+signal-to-noise (CITE THINGS)---is sufficient.
+And yet, the best published \RV\ data sets show empirical scatter
+(rms variations around best-fit kinematic models, for example) at the
+$1\,\mps$ level (for example, CITE THINGS).
+One of the motivating questions for this work is: Is it possible that
+the data-analysis methodologies in use are contributing to this
+empirical variance?
+This is a question in information theory and inference.
 
-While the general idea of measuring a Doppler shift by the relative motions of stellar absorption lines is fairly straightforward, there are considerable subtleties to implementing this, from the choice of reference spectrum or template to the treatment of telluric absorption features.
-Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)} predominantly utilize a cross-correlation method with a custom-built list of spectral lines (see e.g.
-Queloz1995, Baranne1996).
-A lesser-used alternative is cross-correlation or maximum-likelihood fitting using a template spectrum generated from the data (Buchhave, Anglada-Escude). 
-This often involves masking or downweighting of the most telluric-contaminated spectral regions.
+\todo{syntactical reference}
 
-On the other hand, gas-cell-calibrated spectrographs like \HIRES\ commonly adopt a more comprehensive forward-modeling approach. 
-In this framework, the instrumental line spread function, the stellar RV, and other parameters are optimized by modeling each observation as a combination of shifted and instrument-convolved template spectra for the star and gas cell.
-[say a bit more]
+While the general idea of measuring a Doppler shift by the relative
+motions of stellar absorption lines is fairly straightforward, there
+are considerable subtleties to implementing this, from the choice of
+reference spectrum or template to the treatment of telluric absorption
+features.
+Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)}
+predominantly utilize a cross-correlation method with a custom-built
+list of spectral lines (see e.g. Queloz1995, Baranne1996).
+A lesser-used alternative is cross-correlation or maximum-likelihood
+fitting using a template spectrum generated from the data (Buchhave,
+Anglada-Escude).
+This often involves masking or downweighting of the most
+telluric-contaminated spectral regions.
+
+On the other hand, gas-cell-calibrated spectrographs like
+\HIRES\ commonly adopt a more comprehensive forward-modeling approach.
+In this framework, the instrumental line spread function, the stellar
+RV, and other parameters are optimized by modeling each observation as
+a combination of shifted and instrument-convolved template spectra for
+the star and gas cell. [say a bit more]
 

### 2019-08-01 variability/var.tex

@@
 information-theoretic RV bounds, provided that the assumptions hold.
-This methodology opens up the possibility of reducing measurement variances,
-and widening the set of stars available for EPRV study.
+This methodology opens up the possibility improving all past and current
+EPRV measurements, and broadening the set of stars available for EPRV study.
 \end{abstract}
@@
 
+\section{Generalities}
+
+We have shown this and that...
+
+We don't want to fit the spectrum with the same model with which we make the fake data.
+So if we are going to fit the data with a model that is a linear superposition of eigenspectra,
+we will not generate fake data as a linear superposition of eigenspectra.
+
+We are trying to be adversarial and extreme.
+Thus our methods will be stronger than they need to be.
+
+Do we need to define different concepts of RV?
+Stellar center-of-mass RV, and stellar surface RV?
+Do we want to have a pure RV component to the stellar variability in our fake data?
+
+We can turn on and off regularization of the stellar com RV to be generated by a Kepler process,
+or a mixture of Keplers?
+
+\section{Fake data}
+
+We have a baseline fake spectrum, shown in ??.
+
+We permit every line center and depth to vary on a one-dimensional latent $z$.
+
+We permit the line shape to vary on a one-dimensional latent $w$.
+
+We add in a surface RV according to one-dimensional latent $u$?
+
+Note that this basis is not in any sense a linear basis!
+It would be close to linear if the line shifts were small relative to the line widths,
+but they aren't.
+That said, it is still low dimensional even in the linear sense.
+That will be important for what follows.
+
+\section{Method: wobble}
+
+Hello World!
+
+\section{Experiments and results}
+
+Hello World!
+
+\section{Discussion}
+
+Hello World!
+
 \acknowledgements

### 2019-08-08 paper/paper.tex
abstract, formatting, intro start

@@
 
+\addtolength{\textheight}{0.8in}
+\addtolength{\topmargin}{-0.4in}
 \setlength{\parindent}{1.4em} % trust in Hogg
@@
 % context
-Extremely precise stellar radial velocity (\RV) measurements are a
-foundational tool in the field of exoplanets, with hundreds of planets
-discovered by the Doppler method to date and many high-resolution
-spectrographs around the world dedicated to search and
+Extremely precise stellar radial velocity (\RV) measurements
+have been used to to discover hundreds of exoplanets, and
+many new spectrographs are being built for exoplanet search and
 characterization.
@@
 and maximum-likelihood methods.
-We draw out the relationships and equivalencies across these methods.
+We draw out the relationships between these.
 We consider both theoretically generated and data-driven templates.
@@
 We also find that 
-the standard practice of fitting a Gaussian to the cross-correlation
+the currently standard practice of fitting a Gaussian to the cross-correlation
 with a binary mask saturates the \CRLB, so long as the binary mask is
-``correct'' in a certain sense.
+correct in a certain sense.
 As the template or mask becomes incorrect, \RV\ precision suffers.
@@
 the stellar and telluric templates does not cost any significant
-information.
+information, once the number of observational epochs is large.
 That is, data-driven stellar and telluric templates can be correct
-enough to saturate the \CRLB, provided that there are multiple epochs
-of observations and the star does not vary substantially between
-epochs.
+enough to saturate the \CRLB, provided that
+the star does not vary substantially between epochs.
 If the star does vary substantially, a lot more needs to be known
@@
 Fundamentally, the accuracy of radial-velocity mesurement is limited
-by the quality of the spectral models of the targets of the
-spectroscopy.
+jointly by the signal-to-noise of the spectroscopic data and
+the accuracy (in spectral space) of the spectral models.
 In recent decades, however, a new focus has emerged on \emph{extremely
@@
 $1\,\mps$ level (for example, CITE THINGS).
+We hear rumors of various projects delivering empirical scatters for
+the best stars at the $0.2$ to $0.5\,\mps$
+(J.~Brewer, private communication; F.~Pepe, private communication)
+but they aren't yet published.
 One of the motivating questions for this work is: Is it possible that
-the data-analysis methodologies in use are contributing to this
+the data-analysis methodologies currently in use are contributing to this
 empirical variance?
+That is, are we getting the precision we should be, given the data
+quality?
 This is a question in information theory and inference.
 
-\todo{syntactical reference}
+\todo{HOGG: syntactical reference}
 

### 2019-08-08 paper/paper.tex
more intro

@@
 We hear rumors of various projects delivering empirical scatters for
-the best stars at the $0.2$ to $0.5\,\mps$
-(J.~Brewer, private communication; F.~Pepe, private communication)
+the best stars at the $0.2$ to $0.5\,\mps$ level
+(J.~Brewer \& D.~Fischer, private communication; F.~Pepe, private communication)
 but they aren't yet published.
-One of the motivating questions for this work is: Is it possible that
+One of the motivating questions for this \documentname\ is: Is it possible that
 the data-analysis methodologies currently in use are contributing to this
@@
 
+While the general idea of measuring a Doppler shift by the relative
+motions of stellar absorption lines is fairly straightforward, there
+are considerable subtleties.
+These start at the design of the spectrograph hardware
+and include 
+the observing strategy, calibration methodology, spectral extraction
+choice of reference spectrum or template, and the treatment of telluric absorption
+features.
+Here we will focus on the last parts of this: We will presume that the
+spectrograph is delivering wavelength-calibrated, extracted, continuum-normalized
+spectra with minimal issues, and we will consider how to obtain the most
+precise possible \RV\ measurements from those good spectra.
+A more complete treatment would consider the hardware, observing strategy,
+calibration methodology, spectral extraction, and \RV\ measurement as
+an integrated system.
+That's an important direction for future research.
+
 \todo{HOGG: syntactical reference}
 
-While the general idea of measuring a Doppler shift by the relative
-motions of stellar absorption lines is fairly straightforward, there
-are considerable subtleties to implementing this, from the choice of
-reference spectrum or template to the treatment of telluric absorption
-features.
 Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)}

### 2019-08-08 paper/paper.tex
more intro

@@
 motions of stellar absorption lines is fairly straightforward, there
-are considerable subtleties.
-These start at the design of the spectrograph hardware
-and include 
-the observing strategy, calibration methodology, spectral extraction
-choice of reference spectrum or template, and the treatment of telluric absorption
+are considerable complexities and subtleties.
+These start at the design of the spectrograph hardware itself
+and then go on to include 
+the observing strategy, calibration methodology, spectral extraction,
+choice of reference spectrum or template, and treatment of telluric absorption
 features.
@@
 calibration methodology, spectral extraction, and \RV\ measurement as
-an integrated system.
-That's an important direction for future research.
-
-\todo{HOGG: syntactical reference}
+an integrated system of hardware, software, and operations.
+That's an important direction for future research, but out of scope for this \documentname.
 
@@
 A lesser-used alternative is cross-correlation or maximum-likelihood
-fitting using a template spectrum generated from the data (Buchhave,
-Anglada-Escude).
+fitting using a template spectrum generated from the data \todo{(Buchhave,
+Anglada-Escude)}.
 This often involves masking or downweighting of the most
 telluric-contaminated spectral regions.
-
-On the other hand, gas-cell-calibrated spectrographs like
-\HIRES\ commonly adopt a more comprehensive forward-modeling approach.
+As an alternative, gas-cell-calibrated spectrographs like
+\HIRES\ \todo{(CITE)} commonly adopt more of a forward-modeling approach.
 In this framework, the instrumental line spread function, the stellar
@@
 a combination of shifted and instrument-convolved template spectra for
-the star and gas cell. [say a bit more]
+the star and gas cell (CITE).
+Here we focus on the ultra-stabilized spectrograph hardware, but consider
+a few different data-analysis approaches.
+Most of what we do here could be generalized to data generated by gas-cell hardware.
+
+\todo{HOGG: syntactical reference}
 

### 2019-08-08 paper/paper.tex
deletion
tweak

@@
 spectrograph is delivering wavelength-calibrated, extracted, continuum-normalized
-spectra with minimal issues, and we will consider how to obtain the most
+spectra with minimal issues.
+We will consider how to obtain the most
 precise possible \RV\ measurements from those good spectra.
@@
 \todo{HOGG: syntactical reference}
-
-Our primary aim in this work is to establish a common frame of reference for these methods based on their relative ability to extract maximum-precision RVs and their robustness in the face of various expected structured noise sources.
 

### 2019-08-09 paper/paper.tex
intro done

@@
 a few different data-analysis approaches.
-Most of what we do here could be generalized to data generated by gas-cell hardware.
+Most of what we do here could be generalized to data generated by
+gas-cell hardware; we will return to this point in the discussion.
 
-\todo{HOGG: syntactical reference}
-
-For the purposes of this work, we assume that our data set consists of multiple observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts.
-This means we are not considering the case of a real-time data reduction pipeline. 
-We also assume a \HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. 
-We return to this point in the discussion \todo{(do this)}.
+For the purposes of this \documentname, we will assume that the data
+set at hand consists of multiple observations of the same star with
+good coverage throughout the observing season.
+We will assume that the resolution of the spectrograph is $(\lambda /
+\Delta\lambda) > 30,000$, and the star is not near the ecliptic pole,
+so that the observing coverage also delivers significant RV shifts
+between star and atmosphere (and hence between star and tellurics).
+We are not onsidering the case of a real-time data reduction pipeline,
+because such a real-time pipeline does not, at the beginning of an
+observing campaign, have many epochs at hand.
 

### 2019-08-09 paper/paper.tex
more intro

@@
 
-We begin in Section \ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum. 
-In Section \ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting. 
-We then implement these techniques to extract \RV s using 5-\ang-wide sections of simulated and real spectra, as described in Section \ref{s:data}. 
-We compare the relative performances of the methods in Section \ref{s:experiments} and conclude with some recommendations for future RV pipelines in Section \ref{s:summary}.
+We begin in \sectionname~\ref{s:info} with a derivation of the information-theory-based limit on the RV information content of a spectrum. 
+In \sectionname~\ref{s:methods}, we summarize common RV extraction practices used by current \RV\ pipelines, including cross-correlation and template fitting. 
+We then implement these techniques to extract \RV s using 5-\ang-wide sections of simulated and real spectra.
+The simulated and real data are described in \sectionname~\ref{s:data}. 
+We compare the relative performances of the methods in \sectionname~\ref{s:experiments} and conclude with some recommendations for future RV pipelines in \sectionname~\ref{s:summary}.
 
@@
 \label{s:info}
+
+\todo{HOGG: SYNTACTICAL REFERENCE}
 
@@
 Every line was convolved with the same line spread function, which was taken to be a Gaussian with $\sigma$ = 0.05 \ang. 
-In the simplest test case, an arbitrary single line at 5000 \ang\ was used (see Section \ref{s:singleline}). 
+In the simplest test case, an arbitrary single line at 5000 \ang\ was used (see \sectionname~\ref{s:singleline}). 
 However, for most of our tests we injected a set of 12 lines with centers between 4998 - 5002 \ang. 

### 2019-08-09 paper/paper.tex
tweak

@@
 
-\keywords{
-atmospheric effects
+\keywords{\raggedright
+atmospheric~effects
 ---
-binaries: spectroscopic
+binaries:~spectroscopic
 ---
-methods: data analysis
+methods:~data analysis
 ---
-methods: statistical
+methods:~statistical
 ---
-techniques: radial velocities
+techniques:~radial~velocities
 ---
-planets and satellites: detection
+planets~and~satellites:~detection
 }

### 2019-08-15 paper/eprv.tex
abstract tweaks

--- 2019-08-09 paper/paper.tex
@@
 Extremely precise stellar radial velocity (\RV) measurements
-have been used to to discover hundreds of exoplanets, and
+have been used to discover hundreds of exoplanets, and
 many new spectrographs are being built for exoplanet search and
 characterization.
-Current observations are delivering \RV\ measurements at $\mps$
-precision.
-It appears that many hardware systems ought to be---on the basis of
-both spectrograph-stability and information-theoretic
-considerations---capable of delivering even higher precisions.
+The best published results show empirical precision at the $\mps$ level.
+Information theory suggests that it ought to be possible ot reach
+higher precisions with current observations.
 % aims
@@
 We also find that 
-the currently standard practice of fitting a Gaussian to the cross-correlation
+the standard practice of fitting a Gaussian to the cross-correlation
 with a binary mask saturates the \CRLB, so long as the binary mask is
@@
 information, once the number of observational epochs is large.
-That is, data-driven stellar and telluric templates can be correct
-enough to saturate the \CRLB, provided that
+That is, data-driven stellar and telluric templates can be used to
+saturate the \CRLB, provided that
 the star does not vary substantially between epochs.
@@
 
+\clearpage
 \section{Introduction}

### 2019-08-26 paper/eprv.tex
crazy clearpage
information-theory tweaks

@@
 
+\section*{}
 \clearpage
@@
 dynamics to large-scale structure and cosmic expansion.
-Fundamentally, the accuracy of radial-velocity mesurement is limited
+Fundamentally, the accuracy of radial-velocity measurement is limited
 jointly by the signal-to-noise of the spectroscopic data and
@@
 
-\todo{HOGG: SYNTACTICAL REFERENCE}
-
 Given finite, noisy data, there is a limit coming from information
@@
 require a likelihood function.
-As an aside: Bayesians need a likelihood function for inference, and
+As an aside, it is worth noting that
+Bayesians need a likelihood function for inference, and
 frequentists need a likelihood function for analyzing their estimators.
@@
 
-In particular, the \CRLB\ is related to the Fisher Information that is
-frequently used in cosmological experiment forecasts (\todo{CITE
-  SHIT}).
-It is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
-estimator (that is, the information in the velocity estimator) must be smaller than the
+For our purposes, the \CRLB\ is that the \emph{inverse variance} $\sigma_v^{-2}$ of any velocity
+estimator (that is, the information in the velocity estimator) must not be larger than the
 second derivative of the \emph{negative} log likelihood function:
@@
 expectation under the likelihood).
-If there are many other nuisance parameters, a Bayesian can marginalize them
+This expectation value of a second derivative is the Fisher Information
+(information has units of inverse variance).
+This is the information-theory object that is frequently used in
+cosmological experiment forecasts (\todo{CITE SHIT}),
+and it is the quantity used for what we often call the ``formal error'' or
+``formal uncertainty'' on a least-square fit (\todo{CITE fitting a line}).
+The bound is an upper bound on the inverse variance, and thus a lower bound
+on the uncertainty.
+
+If there are many nuisance parameters, a Bayesian can marginalize them
 out to make $L$ the marginalized likelihood for the velocity, or a frequentist
 can include them in the estimation and compute an inverse covariance matrix $C^{-1}$
-for the full parameter vector, which will have the bound
+for the full parameter vector.
+This will then will have the bound
 \begin{equation}
@@
 tensor, corresponding parameters are $\theta_i, \theta_j$, and $v$ is
-any arbitrary vector.  That is, the bound is not just on every
-parameter, but really on on any combination of parameters.
+any arbitrary vector.
+That is, the bound is not just on every parameter, but really on on
+any combination of parameters.
 
@@
 describing the true variance and covariances of the noise process.
+
+\todo{HOGG: SYNTACTICAL REFERENCE}
 

### 2019-08-26 paper/eprv.tex
done with information theory for now

@@
 
-\todo{HOGG: SYNTACTICAL REFERENCE}
-
 Formulae (bounds) related to (\ref{eq:crlb}) have appeared previously
@@
 That work also replaces the model spectrum with a set of optimal
-weights for cross-correlation, which amounts to very much the same thing.
+weights for cross-correlation, which amounts to very much the same thing,
+as we will discuss below.
 Figueira (\todo{cite}) and \todo{others (cite)} show a per-line
@@
 nothing here is controversial.
-We will come back to the scaling of uncertainty with various Gaussian line parameters in \sectionname~\ref{s:singleline}.
+We will come back to the scaling of uncertainty with various
+Gaussian line parameters in \sectionname~\ref{s:singleline}.
 
@@
 \label{s:methods}
+
+\todo{HOGG: SYNTACTICAL REFERENCE}
 

### 2019-08-26 paper/eprv.tex
note to self
clarity

@@
 \end{equation}
-where $L$ is the likelihood, or the probability of the data given the velocity,
+where $L$ is the likelihood, or the probability of the data given the radial velocity $v$,
 and the expectation $E[\cdot]$ is taken over all possible data (that is, it is an
@@
 \begin{equation}
-v^T\cdot C^{-1}\cdot v \leq v^T\cdot Q\cdot v
+u^T\cdot C^{-1}\cdot u \leq u^T\cdot Q\cdot u
 \end{equation}
@@
 where $Q$ is the information tensor, $i,j$ are indices into that
-tensor, corresponding parameters are $\theta_i, \theta_j$, and $v$ is
+tensor, corresponding parameters are $\theta_i, \theta_j$, and $u$ is
 any arbitrary vector.
@@
 \end{equation}
-where the derivatives are of the true spectrum with respect to
-velocity (which can be thought of as the derivative of a Doppler
-operator with respect to velocity acting on the true spectrum), the
+where the derivatives are of the true spectrum $f$ with respect to
+velocity $v$ (which can, in turn, be thought of as the derivative of a Doppler
+operator with respect to velocity acting on the true spectrum; CITE WOBBLE), the
 derivatives are column vectors, and $C^{-1}$ is the inverse of the covariance matrix
@@
 \todo{HOGG: SYNTACTICAL REFERENCE}
+
+HOGG: OPEN by giving the LF and what it's optimum is, in the simplest case.
+Parameterize $f(v)$ where $v$ is the \RV.
 

### 2019-08-26 paper/eprv.tex
cross correlation

@@
 \begin{equation}\label{eq:crlb}
-\frac{1}{\sigma_v^2} \leq \left[\frac{\dd f}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f}{\dd v}\right]
+\frac{1}{\sigma_v^2} \leq \left[\frac{\dd f(v)}{\dd v}\right]\T\cdot C^{-1}\cdot\left[\frac{\dd f(v)}{\dd v}\right]
 \end{equation}
-where the derivatives are of the true spectrum $f$ with respect to
-velocity $v$ (which can, in turn, be thought of as the derivative of a Doppler
+where the derivatives are of the true spectrum $f(v)$ with respect to
+radial velocity $v$ (which can, in turn, be thought of as the derivative of a Doppler
 operator with respect to velocity acting on the true spectrum; CITE WOBBLE), the
@@
 
-\todo{HOGG: SYNTACTICAL REFERENCE}
+\CRLB-saturating estimators are (or are equivalent to)
+maximum-likelihood estimators.
+When the likelihood is Gaussian, the log-likelihood
+is a sum of a quadratic chi-squared-like term and a determinant term
+\begin{equation}
+\ln L = -\frac{1}{2}\,[y - f(v)]\T\cdot C^{-1}\cdot [y - f(v)] - \frac{1}{2}\,||2\pi\,C|| \quad ,
+\label{eq:lf}
+\end{equation}
+where $y$ is a column vector of the data, $f(v)$ is the model spectrum,
+which depends on the \RV\ $v$,
+and $C$ is the covariance matrix describing the noise variance.
+In the case of independent pixel noise the chi-squared term just becomes an
+inverse-variance weighted sum of squared residuals (chi-squared).
+The maximum-likelihood estimator for $v$ is the value of $v$ which maximizes
+the log-likelihood.
+Since only the chi-squared term depends on $v$, maximizing the likelihood
+means minimizing chi-squared.
 
-HOGG: OPEN by giving the LF and what it's optimum is, in the simplest case.
-Parameterize $f(v)$ where $v$ is the \RV.
-
-Most methods for radial-velocity estimation involve some kind of
+Most methods for radial-velocity estimation aren't explicitly maximizing
+any likelihood.
+Instead, they involve some kind of
 cross-correlation of the data with a template.
@@
 estimators will be maximum-likelihood estimators.
-However, under some simple assumptions, cross-correlation optima are
-also maximum-likelihood optima.
-
-For example, when the likelihood is Gaussian, the log-likelihood
-contains a chi-squared-like term
+However, under simple assumptions, cross-correlation optima are
+also maximum-likelihood optima. The argument is as follows:
+If we expand the quadratic form (chi-squared term) in the
+likelihood in equation~(\ref{eq:lf}), it becomes
 \begin{equation}
--\frac{1}{2}\,[y - f]\T\cdot C^{-1}\cdot [y - f] \quad ,
-\end{equation}
-where $y$ is a column vector of the data, $f$ is the model spectrum,
-and $C$ is the covariance matrix describing the noise variance.
-In the case of independent pixel noise this just becomes an
-inverse-variance weighted sum of squared residuals (chi-squared).
-If we expand this quadratic form, it becomes
-\begin{equation}
-{}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f 
--\frac{1}{2}\,f\T\cdot C^{-1}\cdot f \quad.
+{}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f(v) 
+-\frac{1}{2}\,f(v)\T\cdot C^{-1}\cdot f(v) \quad.
 \end{equation}
@@
 cross-correlation.
-For this reason, we will use a variant of cross-correlation
-optimization as our radial-velocity estimator.
+For this reason, we can use a variant of cross-correlation
+optimization as a radial-velocity estimator.
+
+\todo{HOGG: SYNTACTICAL REFERENCE}
 

### 2019-08-26 paper/eprv.tex
tweaks

@@
 When the likelihood is Gaussian, the log-likelihood
-is a sum of a quadratic chi-squared-like term and a determinant term
+is a sum of a quadratic chi-squared-like term and a log-determinant term
 \begin{equation}
-\ln L = -\frac{1}{2}\,[y - f(v)]\T\cdot C^{-1}\cdot [y - f(v)] - \frac{1}{2}\,||2\pi\,C|| \quad ,
+\ln L = -\frac{1}{2}\,[y - f(v)]\T\cdot C^{-1}\cdot [y - f(v)] - \frac{1}{2}\,\ln||2\pi\,C|| \quad ,
 \label{eq:lf}
@@
 That is, the chi-squared term factors into a term that is the square
-of the model, a term that is the square of the data, and a term that
-is a cross-correlation.
-As long as the amplitude of $f$ doesn't depend on velocity, and the
-$C$ tensor is ``smooth'' in certain respects, the dependence of this
+of the data $y$, a term that is the square of the model $f(v)$,
+and a term that is a cross-correlation of data times model.
+As long as the amplitude of $f(v)$ doesn't depend strongly on velocity $v$,
+and the $C$ tensor is ``smooth'' in certain respects, the dependence of this
 chi-squared term on velocity will be dominated by the

### 2019-08-28 paper/eprv.tex
cross correlations orama

@@
 \begin{equation}
-{}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y + y^T\cdot C^{-1}\cdot f(v) 
+ {}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y
+               + y\T\cdot C^{-1}\cdot f(v)
 -\frac{1}{2}\,f(v)\T\cdot C^{-1}\cdot f(v) \quad.
@@
 of the data $y$, a term that is the square of the model $f(v)$,
-and a term that is a cross-correlation of data times model.
+and a term that is a scalar product or projection or
+cross correlation of data with model.
 As long as the amplitude of $f(v)$ doesn't depend strongly on velocity $v$,
@@
 For this reason, we can use a variant of cross-correlation
-optimization as a radial-velocity estimator.
+optimization as a radial-velocity estimator, or, in other words, the
+\RV\ $v$ that optimizes $y\T\cdot C^{-1}\cdot f(v)$ will also (nearly)
+optimize the likelihood.
+
+In actually performing this cross-correlation optimization,
+investigators make some non-trivial choices.
+The first is whether or not to use the inverse variance tensor $C$.
+Many codes do not make any use of the uncertainty estimates at all in
+the calculation of the cross correlation.
+Not using the uncertainties (treating $C$ as the identity) is
+equivalent to assuming that the pixel uncertainties don't vary much
+from pixel to pixel across the spectrum.
+If the cross correlation is optimized with $C$ set to the identity,
+the optimum of that doesn't strictly optimize the likelihood function;
+the differences depdends on the variations in the uncertainties.
+Another choice is how finely to sample in radial velocity $v$, or how
+to perform the optimization.
+If the $v$ spacing is coarse (but not too coarse!), quadratic
+interpolation can be used to find the optimum at high precision.
+The velocity spacing need only be smaller than something like half the
+spectral velocity resolution ($c\,\Delta\lambda / \lambda$).
+This problem is similar to stellar centroiding methods in use in
+imaging surveys (CITE VAKILI).
+Another choice faced by an investigator is what to do at the ends or
+edges of the spectra or spectral segments used for the
+cross-correlation.
+If the edges are treated naively, it is possible to introduce unintended
+dependencies that invalidate the requirement that the model-times-model term
+(the $f(v)\T\cdot C^{-1}\cdot f(v)$ term) not depend on \RV\ $v$.
+
+HOGG ...In detail, in the experiments that follow, the
+cross-correlation we will optimize has the following properties...
+
+The maximum-likelihood and cross-correlation methods only give
+bound-saturating \RV\ estimates when the spectral model $f(v)$
+respresents something similar to the true expecation of the spectrum
+at velocity $v$.
+And yet there is a paradox, because many world-class \RV\ pipelines
+(like for instance HOGG CITE) perform cross-correlation with a
+\emph{binary mask}.
+This is not a good representation of the spectral expectation! So why
+do these methods work to deliver high quality \RV\ estimates.
+The answer is that the codes that cross-correlate with a binary mask
+subsequently fit that cross-correlation with a Gaussian function.
+It turns out that the operation ``cross-correlate the data with a
+binary mask, followed by fitting a Gaussian to the cross-correlation
+function'' is mathematically equivalent to ``smooth the binary mask
+with a Gaussian, and then cross-correlate with the data''.
+That is, when a pipeline cross-correlates with a binary mask and then
+fits that cross-correlation function with a Gaussian, it is equivalent
+to smoothing the binary mask with a best-fitting Gaussian function to
+make a smooth, realistic $f(v)$ and then cross-correlating with that
+realistic $f(v)$.
+The fitting gives the flexibility to match the spectral resolution of
+the template to the spectral resolution of the data, in the
+approximation that the resolution is close to constant over the
+spectral range in velocity units.
+So in all cases of bound-saturating \RV\ measurements, the
+\RV\ estimation is effectively a good approximation to
+maximum-likelihood with a good spectral template.
 
 \todo{HOGG: SYNTACTICAL REFERENCE}
-
-...In detail, the cross-correlation we will optimize is...
 

### 2019-08-29 paper/eprv.tex
tweak
estimators section done

@@
 \newcommand{\sectionname}{Section}
+\renewcommand{\paragraph}[1]{\bigskip\par\noindent\textsl{#1} ---}
 \newcommand{\todo}[1]{\textcolor{red}{#1}}  % gotta have \usepackage{xcolor} in main doc or this won't work
+\newcommand{\CITE}{\todo{CITE}}
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 been put into the challenges of making measurements at this level of
-precision (CITE some reviews).
+precision (\CITE some reviews).
 It appears that contemporary spectrographs are stable at the
-$0.05\,\mps$ level (CITE THINGS), and that the information content in
+$0.05\,\mps$ level (\CITE THINGS), and that the information content in
 the data---which is set by the spectral features, the spectrograph
 resolution, the wavelength range, and the observational
-signal-to-noise (CITE THINGS)---is sufficient.
+signal-to-noise (\CITE THINGS)---is sufficient.
 And yet, the best published \RV\ data sets show empirical scatter
 (rms variations around best-fit kinematic models, for example) at the
-$1\,\mps$ level (for example, CITE THINGS).
+$1\,\mps$ level (for example, \CITE THINGS).
 We hear rumors of various projects delivering empirical scatters for
@@
 
-Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& cite)}
+Ultra-stabilized spectrographs like \HARPS\ \todo{(define \& \CITE)}
 predominantly utilize a cross-correlation method with a custom-built
@@
 a combination of shifted and instrument-convolved template spectra for
-the star and gas cell (CITE).
+the star and gas cell (\CITE).
 Here we focus on the ultra-stabilized spectrograph hardware, but consider
@@
 radial velocity $v$ (which can, in turn, be thought of as the derivative of a Doppler
-operator with respect to velocity acting on the true spectrum; CITE WOBBLE), the
+operator with respect to velocity acting on the true spectrum; \CITE WOBBLE), the
 derivatives are column vectors, and $C^{-1}$ is the inverse of the covariance matrix
@@
 as we will discuss below.
-Figueira (\todo{cite}) and \todo{others (cite)} show a per-line
+Figueira (\todo{cite}) and \todo{others (\CITE)} show a per-line
 bound, based on assumptions similar to those of Bouchy but with the additional assumption that
@@
 
+There are three dominant methods for the estimation of a radial veloctity,
+given spectral data.
+
+\paragraph{Maximum-likelihood estimator}
 \CRLB-saturating estimators are (or are equivalent to)
@@
 \begin{equation}
-\ln L = -\frac{1}{2}\,[y - f(v)]\T\cdot C^{-1}\cdot [y - f(v)] - \frac{1}{2}\,\ln||2\pi\,C|| \quad ,
+\ln L = -\frac{1}{2}\,[y - f(v)]\T\cdot C^{-1}\cdot [y - f(v)] - \frac{1}{2}\,\ln||2\pi\,C||
+\quad ,
 \label{eq:lf}
@@
 means minimizing chi-squared.
+Optimization is straightforward; it is one-dimensional optimization
+(given the template spectrum $f(v)$, and derivatives with respect to
+velocity are easy to take:
+\begin{equation}
+\frac{\dd\ln L}{\dd v} = [\frac{\dd f(v)}{\dd v}]\T\cdot C^{-1}\cdot [y - f(v)]
+\quad .
+\label{eq:dldv}
+\end{equation}
 
+\paragraph{Cross correlation with a template}
 Most methods for radial-velocity estimation aren't explicitly maximizing
@@
  {}-\frac{1}{2}\,y\T\cdot C^{-1}\cdot y
-               + y\T\cdot C^{-1}\cdot f(v)
+            + f(v)\T\cdot C^{-1}\cdot y
 -\frac{1}{2}\,f(v)\T\cdot C^{-1}\cdot f(v) \quad.
@@
 optimization as a radial-velocity estimator, or, in other words, the
-\RV\ $v$ that optimizes $y\T\cdot C^{-1}\cdot f(v)$ will also (nearly)
+\RV\ $v$ that optimizes $f(v)\T\cdot C^{-1}\cdot y$ will also (nearly)
 optimize the likelihood.
+
+Another argument is as follows:
+As long as the derivative $\dd f(v)/\dd v$ is close to orthogonal
+(given the $C^{-1}$ metric) to the template $f(v)$, the derivative of
+the log likelihood given in equation~(\ref{eq:dldv}) is equal to the
+derivative of the cross-correlation $f(v)\T\cdot C^{-1}\cdot y$.
+Therefore, whatever optimizes the likelihood will (nearly) optimize
+the cross-correlation.
 
@@
 This problem is similar to stellar centroiding methods in use in
-imaging surveys (CITE VAKILI).
+imaging surveys (\CITE VAKILI).
 Another choice faced by an investigator is what to do at the ends or
@@
 
-HOGG ...In detail, in the experiments that follow, the
-cross-correlation we will optimize has the following properties...
+\todo{HOGG ...In detail, in the experiments that follow, the
+cross-correlation we will optimize has the following properties...}
 
+\paragraph{Cross correlation with a binary mask}
 The maximum-likelihood and cross-correlation methods only give
@@
 And yet there is a paradox, because many world-class \RV\ pipelines
-(like for instance HOGG CITE) perform cross-correlation with a
+(like for instance \\CITE) perform cross-correlation with a
 \emph{binary mask}.
 This is not a good representation of the spectral expectation! So why
-do these methods work to deliver high quality \RV\ estimates.
+do these methods deliver high quality \RV\ estimates?
 The answer is that the codes that cross-correlate with a binary mask
@@
 
-\todo{HOGG: SYNTACTICAL REFERENCE}
-
-A variety of choices exist for the cross-correlation template $f$. A few possibilities include:
+\paragraph{Choice of template}
+A variety of choices exist for the cross-correlation template
+$f(v)$. A few possibilities include:
 \begin{itemize}
-\item a ``binary'' mask, or a spectrum of zero value with non-zero peaks at the expected rest-frame wavelengths of the spectral lines of interest.
-A cross-correlation between this mask and the observed spectrum will effectively compute the average line profile with its minimum corresponding to the best-fit RV.
-This method has the benefit of trivially excluding telluric-contaminated or otherwise unreliable regions of the spectrum by deleting them from the mask.
-A desired relative weighting of lines can be incorporated by setting the non-zero mask values to different amplitudes at different lines (cite Pepe).
-This method is currently used by the \HARPS\ pipeline (cite).
-\item synthetic spectrum (i.e.
-a slightly wrong model prediction, since the stellar model will never be a perfect reproduction of the star in question): sub-tweet: some surveys cross-correlate with a single (same) template for every single star, which is extremely wrong.
-you can be more right by using a spectral-type-specific template.
-\item data-driven template spectrum: some surveys (HIRES) use a single observation as a template. combining information from many visits is better because the combined SNR is amazing.
+\item \emph{A theoretically produced synthetic spectrum:}
+  This will be a slightly wrong model prediction,
+  since no stellar model will produce perfect reproduction of the
+  spectral expectation for the star in question.
+  The chosen template should be matched as closely as possible however
+  to the target star. Some surveys cross correlate with one universal
+  template for all stars; this will lead to increases in both bias and
+  variance relative to the \CRLB.
+\item \emph{A binary mask:}
+  This is a or a spectrum of zero values with non-zero top-hats at the
+  expected rest-frame wavelengths of the spectral lines of interest.
+  A cross-correlation between this mask and the observed spectrum will
+  effectively compute the average line profile with its minimum
+  corresponding to the best-fit RV.
+  As we note above, fitting that mean line profile with a Gaussian
+  before identifying the \RV\ estimate is equivalent to smoothing the
+  mask with a Gaussian and then using it as the stellar template in a
+  straightforward cross correlation.
+  The binary-mask method has the benefit of exactly zeroing out
+  telluric-contaminated or otherwise unreliable regions of the
+  spectrum (by deleting them from the mask).
+  A desired relative weighting of lines can be incorporated by setting
+  the non-zero mask values to different amplitudes at different lines
+  (\CITE Pepe). This method is currently used by the \HARPS\ pipeline
+  (\CITE), and an example is shown in \figurename~\ref{fig:binarymask}.
+\item \emph{A data-driven estimate of the template spectrum:}
+  Some surveys (for example, \HIRES, \CITE) use a single observation as a
+  template.
+  Combining information from many visits is better because the
+  combined SNR will be higher.
+  In what follows, we show that with sufficient epochs, a data-driven
+  template is accurate enough to saturate the \CRLB.
 \end{itemize}
-
-One final implementation detail is the following:
-When we optimize the cross-correlation function, we don't run an optimizer;
-we perform brute-force search on a linear velocity grid.
-This grid, however, does not need to be fine.
-We use a coarse grid (\todo{give grid spacing here}), which does not
-over-resolve the velocity peak at the optimum.
-We then take the highest value on the grid and the values on either
-side, fit a parabola through those three values, and take as the optimal
-velocity the peak of that parabola.
-That is, we obtain sub-grid velocity resolution by means of a quadratic
-interpolation at the peak on the finite grid.
-\todo{Do we need to demonstrate this with a toy figure?}
-
-\begin{figure}
+\begin{figure}[htp]
 \centering
@@
 \label{s:data}
+
+\todo{HOGG SYNTACTICAL}
 

### 2019-08-29 paper/eprv.tex
typo

@@
 \item \emph{A binary mask:}
-  This is a or a spectrum of zero values with non-zero top-hats at the
+  This is a pseudo-spectrum of zero values with non-zero top-hats at the
   expected rest-frame wavelengths of the spectral lines of interest.

### 2021-06-11 paper/eprv.tex
affiliation tweak; testing sync
notes 2 self

@@
 
-\documentclass[modern]{aastex62}
+\documentclass[modern]{aastex63}
 \usepackage{graphicx}
@@
 \author[0000-0001-9907-7742]{Megan Bedell}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
-\affil{Department of Astronomy \& Astrophysics, University of Chicago, 5640 S. Ellis Ave, Chicago, IL 60637, USA}
+\affil{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 
 \author[0000-0003-2866-9403]{David W. Hogg}
-\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
+\affil{Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
 \affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
@@
 \end{equation}
+\todo{HOGG CHECK THIS (factor of 2?)}
 where $L$ is the likelihood, or the probability of the data given the radial velocity $v$,
@@
 This expectation value of a second derivative is the Fisher Information
-(information has units of inverse variance).
+(information, for us, has units of inverse variance).
 This is the information-theory object that is frequently used in
-cosmological experiment forecasts (\todo{CITE SHIT}),
+cosmological experiment forecasts (\todo{CITE}),
 and it is the quantity used for what we often call the ``formal error'' or
@@
 \end{itemize}
-It is worthy of note that every formula (that we know) in the
-literature for maximum \RV\ precision (\todo{cite Butler, Bouchy,
-  Figueira, Lovis}) makes all of these same assumptions,
-plus additional \emph{even more restrictive assumptions}.
-Under these disturbingly unrealistic assumptions, the \CRLB\ becomes
+Under these unrealistic assumptions, the \CRLB\ becomes
 \begin{equation}\label{eq:crlb}
@@
 
-Formulae (bounds) related to (\ref{eq:crlb}) have appeared previously
-in the literature, although in somewhat more complicated forms.
+It is worthy of note that every formula (that we know) in the
+literature, comparable to (\ref{eq:crlb}), for maximum \RV\ precision (\todo{cite Butler, Bouchy,
+  Figueira, Lovis}) makes all of these same assumptions,
+plus additional \emph{even more restrictive assumptions}.
 Butler (\todo{cite}) shows a similar expression, but makes the additional

### 2021-06-11 paper/eprv.tex
remove wrongish par

@@
 
-Another argument is as follows:
-As long as the derivative $\dd f(v)/\dd v$ is close to orthogonal
-(given the $C^{-1}$ metric) to the template $f(v)$, the derivative of
-the log likelihood given in equation~(\ref{eq:dldv}) is equal to the
-derivative of the cross-correlation $f(v)\T\cdot C^{-1}\cdot y$.
-Therefore, whatever optimizes the likelihood will (nearly) optimize
-the cross-correlation.
-
 In actually performing this cross-correlation optimization,
 investigators make some non-trivial choices.
-The first is whether or not to use the inverse variance tensor $C$.
-Many codes do not make any use of the uncertainty estimates at all in
+The first is whether or not to use the inverse variance tensor $C^{-1}$.
+Many codes do not make any use of uncertainty estimates at all in
 the calculation of the cross correlation.
@@
 And yet there is a paradox, because many world-class \RV\ pipelines
-(like for instance \\CITE) perform cross-correlation with a
+(like for instance \CITE) perform cross-correlation with a
 \emph{binary mask}.

### 2021-06-12 paper/eprv.tex
up to 6.3.1

@@
 
-\documentclass[modern]{aastex63}
+\documentclass[modern]{aastex631}
 \usepackage{graphicx}
@@
 
-\acknowledgements
+\begin{acknowledgements}
 It is a pleasure to thank
@@
 \todo{...add software tags}
+\end{acknowledgements}
 

### 2021-06-12 paper/eprv.tex
Updates from Overleaf

--- 2021-06-11 paper/eprv.tex
@@
 % -----------
+% - update aastex and keywords
 % - search for all \todo
@@
 \keywords{\raggedright
+\todo{there is a new keyword set; these are wrong now:}
 atmospheric~effects
